---
title: "Model Comparison and Averaging"


bibliography: stat-mod.bib
csl: conservation-biology.csl

---

```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false

library(knitr)
library(ggplot2)
library(cowplot)
library(dplyr)
library(MASS)

opts_chunk$set(echo = FALSE)

# shared data setup used throughout
clim <- read.csv("data/plot_climate.csv")
geo_hii <- read.csv("data/hii_geo.csv")
tree <- read.csv("data/OpenNahele_Tree_Data.csv")

all_dat <- left_join(tree, clim, by = "PlotID") |>
    left_join(dplyr::select(geo_hii, !(lon:lat)), by = "PlotID")

dat_plt <- group_by(all_dat, PlotID) |>
    summarize(across(c(Island, Plot_Area, lon:elev_m), first),
              nspp = n_distinct(Scientific_name)) |>
    ungroup() |>
    filter(Plot_Area < 10000)
```


## What predicts species richness best?

***Temperature*** or ***Human Impact***?



::: {.fragment}

We saw that including both in one model had problem of collinearity

:::

::: {.fragment}

But does one predict richness better than another?

:::


::: {.fragment}

We cannot use LRT because these are not nested models, in fact they have the same d.f.

::: {.columns}

:::: {.column}

$$
\begin{aligned}
\log(\mu_\text{spp richness}) &= b_0 + b_1 \log(\text{area}) + b_2 \text{temperature} \\
\log(\mu_\text{spp richness}) &= b_0 + b_1 \log(\text{area}) + b_2 (\text{human impact})
\end{aligned}
$$

::::

:::

:::

## This is a job for AIC

***Akaike Information Criterion***


$$
AIC = -2 \mathcal{l}(\hat{\theta}_{MLE}) + 2k
$$

- $\mathcal{l}(\hat{\theta}_{MLE})$ is max of the log likelihood
- $k$ is number of parameters

::: {.fragment}

We can compare non-nested models with AIC

:::


## This is a job for AIC

$$
AIC = -2 \mathcal{l}(\hat{\theta}_{MLE}) + 2k
$$



```{r}
#| echo: true

mod_temp <- glm.nb(nspp ~ log(Plot_Area) + avg_temp_annual_c, 
                   data = dat_plt)
AIC(mod_temp)

mod_hii <- glm.nb(nspp ~ log(Plot_Area) + hii, 
                  data = dat_plt)
AIC(mod_hii)

```

The human impact model is better supported (lower AIC)


## This is a job for AIC

$$
AIC = -2 \mathcal{l}(\hat{\theta}_{MLE}) + 2k
$$

Purpose of $2k$ term is more clear for models with unequal d.f.

```{r}
#| echo: true

mod_temp_precip <- glm.nb(
    nspp ~ log(Plot_Area) + avg_temp_annual_c + rain_annual_mm, 
    data = dat_plt
)
AIC(mod_temp_precip)

AIC(mod_hii)
```


Now the `temp + rain` model is better supported 

## AIC penalizes more parameters

But...

```{r}
#| echo: true

logLik(mod_temp_precip)
logLik(mod_temp)
```

adding parameters will always increase log likelihood

::: {.fragment}

Is the increase real? Or just *"over fitting"*

:::


::: {.fragment}

*"over fitting"* is a term for what we know already: more parameters could let a model wiggle to fit noise

:::

## AIC penalizes more parameters

How? and why?


::: {.r-stack}

:::: {.fragment .fade-in-then-out}

Imagine you have one data set <span style="display:inline-block; width:0.8em; height:0.8em; background-color:black; border-radius:50%; vertical-align:middle;"></span> 

::: {.columns}

:::: {.column}

```{r}
#| fig-width: 6
#| fig-height: 6

set.seed(123)
x <- runif(40)
y <- 3 + x + rnorm(length(x), sd = 0.25)
g <- rep(c("old", "new"), each = 20)

dat <- data.frame(x, y, g)
gp <- ggplot(dat, aes(x, y, color = g)) +
    geom_point(size = 10) +
    geom_smooth(method = "lm", se = FALSE, 
                linewidth = 3, color = "black") +
    theme_cowplot() +
    theme(legend.position = "none", 
          axis.text = element_blank(), 
          axis.title = element_text(size = 34)) +
    scale_color_discrete(palette = c("transparent", "black"))

gp
```

::::

:::: {.column}

and fit a model to it by maximizing likelihood.

::::

:::

::::

:::: {.fragment .fade-in-then-out}

Imagine you have one data set <span style="display:inline-block; width:0.8em; height:0.8em; background-color:black; border-radius:50%; vertical-align:middle;"></span> then collect a new one <span style="display:inline-block; width:0.8em; height:0.8em; background-color:gray; border-radius:50%; vertical-align:middle;"></span>


::: {.columns}

:::: {.column}

```{r}
#| fig-width: 6
#| fig-height: 6

gp +
    geom_point(data = filter(dat, g == "new"), 
               aes(x, y), 
               color = "gray", size = 10)

```

::::

:::: {.column}

and calculate the log likelihood of the new data given the same model fit to the first data



::::

:::


::::

:::

## AIC penalizes more parameters

How? and why?


::: {.r-stack}

:::: {.fragment .fade-in-then-out}

Imagine you have one data set <span style="display:inline-block; width:0.8em; height:0.8em; background-color:black; border-radius:50%; vertical-align:middle;"></span> 

::: {.columns}

:::: {.column}

```{r}
#| fig-width: 6
#| fig-height: 6

set.seed(123)
x <- runif(40)
y <- 3 + x + rnorm(length(x), sd = 0.25)
g <- rep(c("old", "new"), each = 20)

dat <- data.frame(x, y, g)
gp <- ggplot(dat, aes(x, y, color = g)) +
    geom_point(size = 10) +
    geom_smooth(method = "lm", se = FALSE, 
                linewidth = 3, color = "black") +
    theme_cowplot() +
    theme(legend.position = "none", 
          axis.text = element_blank(), 
          axis.title = element_text(size = 34)) +
    scale_color_discrete(palette = c("transparent", "black"))

gp
```

::::

:::: {.column}

and fit a model to it by maximizing likelihood.

::::

:::

::::

:::

## AIC penalizes more parameters

How? and why?


Imagine you have one data set <span style="display:inline-block; width:0.8em; height:0.8em; background-color:black; border-radius:50%; vertical-align:middle;"></span> then collect a new one <span style="display:inline-block; width:0.8em; height:0.8em; background-color:gray; border-radius:50%; vertical-align:middle;"></span>


::: {.columns}

:::: {.column}

```{r}
#| fig-width: 6
#| fig-height: 6

gp +
    geom_point(data = filter(dat, g == "new"), 
               aes(x, y), 
               color = "gray", size = 10)

```

::::

:::: {.column}


::: {.r-stack}

:::: {.fragment .fade-in-then-out}

We could calculate the log likelihood for the new data ***but*** with parameter estimates from original model 

::::

:::: {.fragment}

The difference between the mean log likelihood of the original data and the mean log likelihood of the new data will be about $k$


::::


:::


::::

:::


::: {.absolute top="70%" left="35%" style="z-index: 10;"}

::: {.fragment}

$$
\frac{1}{n} \mathcal{l}(\hat{\theta} \mid \text{data}_\text{orig}) - \frac{1}{n} \mathcal{l}(\hat{\theta} \mid \text{data}_\text{new}) \approx k
$$

:::

:::

## AIC penalizes more parameters

<br><br>

So AIC is trying to account for the chance that additional parameters are just fitting noise by including a penalty for those additional parameters


## Returning to model comparisons

We can rest assured that the improvement in AIC for the `temp + rain` model is not due just to adding antother parameter

```{r}
#| echo: true

AIC(mod_temp_precip)

AIC(mod_hii)
```

::: {.fragment}

A common rule is that models with AIC values within 2 points are effectively equivalent; $>2$ means one model really is better supported [@burnham-anderson]

:::


## Model averaging

::: {.incremental}

- model comparisons tells us which model is more supported, a way of testing hypotheses
- model averaging is less about testing hypotheses and more about ***prediction***
- for example: how can we best predict tree richness across the pae ʻāina?

:::


## Model averaging

Consider the 3 models we have made so far

```{r}
#| echo: true
#| eval: false

nspp ~ log(Plot_Area) + hii
nspp ~ log(Plot_Area) + avg_temp_annual_c
nspp ~ log(Plot_Area) + avg_temp_annual_c + rain_annual_mm
```


::: {.fragment}

Each model makes a different prediction about the expected number of species in a plot

:::

::: {.fragment}

How could we combine the models to make one prediction?

:::

::: {.fragment}

Why do we want to?

:::


## Model averaging

Consider the 3 models we have made so far

```{r}
#| echo: true
#| eval: false

nspp ~ log(Plot_Area) + hii
nspp ~ log(Plot_Area) + avg_temp_annual_c
nspp ~ log(Plot_Area) + avg_temp_annual_c + rain_annual_mm
```

The predictions from the combined model will be more accurate than any one model by itself

::: {.fragment}

And also better than one big model

```{r}
#| echo: true
#| eval: false

nspp ~ log(Plot_Area) + hii + avg_temp_annual_c + rain_annual_mm
```

Because we know, e.g., there are issues of collinearity. *For smaller data sets over fitting one big model is also an issue*

:::


## Model averaging

How do we combine models to make one prediction?


::: {.fragment}

We could average the parameters of each model to make an ***average*** model

$$
\begin{aligned}
\log(\mu_\text{nspp}) = &b_0^{\text(avg.)} + b_1^{\text(avg.)} \log(\text{area}) + b_2^{\text(avg.)} \text{hii} + \\ 
&b_3^{\text(avg.)} \text{temp} + b_4^{\text(avg.)} \text{rain}
\end{aligned}
$$

where, e.g., $b_3^{\text(avg.)}$ is the *average* of all slope estimates for `avg_temp_annual_c` across all the models

:::


## Model averaging

Let's make that more concrete with our actual models. 

::: {.fragment}

Here are the slope estimates for temperature in every model where it shows up


```{r}
#| echo: true


coefficients(mod_temp)["avg_temp_annual_c"]
coefficients(mod_temp_precip)["avg_temp_annual_c"]

```

:::

::: {.fragment}

```{r}
b31 <- coefficients(mod_temp)["avg_temp_annual_c"]
b32 <- coefficients(mod_temp_precip)["avg_temp_annual_c"]
b3avg <- mean(c(b31, b32))
```

So the average slope is $b_3^{\text(avg.)} = \frac{`r round(b31, 3)` + `r round(b32, 3)`}{2} = `r round(b3avg, 3)`$

:::


## Model averaging


Note: there are two ways to average

::: {.columns}

:::: {.column}

average over ***only models*** where the variable shows up
<br>

$b_3^{\text(avg.)} = \frac{`r round(b31, 3)` + `r round(b32, 3)`}{2}$

::::

:::: {.column}

average over ***all models*** no matter if the variable shows up
<br>

$b_3^{\text(avg.)} = \frac{0 + `r round(b31, 3)` + `r round(b32, 3)`}{3}$

::::

:::

<br>

::: {.fragment}

The default in the code we'll use for model average is to use ***only models*** where the variable shows up

:::

## Model averaging

But this simple average parameter value ignores the fact that some individual models are already better at making predictions

::: {.fragment}

The AIC values tell us about the support of each model

```{r}
#| echo: true

AIC(mod_hii)
AIC(mod_temp)
AIC(mod_temp_precip)
```

:::

We want to account for these differences in support when averaging


## Model averaging

We turn AIC values into weights. For model $i$:


$$
w_i = \frac{e^{-\Delta\text{AIC}_i / 2}}{\sum_j e^{-\Delta\text{AIC}_j / 2}}
$$

where $\Delta\text{AIC}_i = \text{AIC}(\text{model}_i) - (\text{minimum AIC})$

::: {.fragment}

The weights sum to 1, they can be roughly interpreted as the probability that $\text{model}_i$ is the best supported model

:::

## Model averaging

Again, let's make this concrete

::: {.columns}

:::: {.column width=33%}

$\text{AIC}_i$

```{r}
#| echo: true

AIC(mod_hii)
AIC(mod_temp)
AIC(mod_temp_precip)
```

::::

:::: {.column width=33%}

$\Delta\text{AIC}_i$

<div style="height: 15px;"></div>

```{r}
#| output: asis

cat(round(AIC(mod_hii) - AIC(mod_temp_precip), 3))
```

<div style="height: 5px;"></div>

```{r}
#| output: asis
cat(round(AIC(mod_temp) - AIC(mod_temp_precip), 3))
```

<div style="height: 5px;"></div>

```{r}
#| output: asis
cat(round(AIC(mod_temp_precip) - AIC(mod_temp_precip), 3))
```

::::


:::: {.column width=33%}

$w_i$

```{r}
wfun <- function(all_mod, i) {
    all_aic <- sapply(all_mod, AIC)
    dAIC <- all_aic - min(all_aic)
    
    e <- exp(-dAIC / 2)
    
    e[i] / sum(e)
}
```


<div style="height: 15px;"></div>

```{r}
#| output: asis

cat(round(wfun(list(mod_hii, mod_temp, mod_temp_precip), 1), 5))
```

<div style="height: 5px;"></div>

```{r}
#| output: asis
round(wfun(list(mod_hii, mod_temp, mod_temp_precip), 2), 5) |> 
    format(scientific = FALSE) |> 
    cat()
```

<div style="height: 5px;"></div>

```{r}
#| output: asis
cat(round(wfun(list(mod_hii, mod_temp, mod_temp_precip), 3), 5))
```

::::


:::

## Package *MuMIn* does this for us!

Create the averaged model

```{r}
#| echo: true

library(MuMIn)

# pass in a list of all models we want to consider
mod_avg <- model.avg(list(mod_hii, mod_temp, mod_temp_precip))

```


::: {.fragment}

Have a look at the averaged coefficients

```{r}
#| echo: true

coefficients(mod_avg)
```

:::

## Package *MuMIn* does this for us!

Look at a similar table of AIC, $\Delta$AIC, and weights


```{r}
#| echo: true

mod_avg$msTable
```

A few things to note:

- order is ranked best supported model to least supported
- AICc is used instead of AIC

::: {.fragment}
AICc is a sample size corrected version AIC, not a bad idea to use
:::


## Package *MuMIn* does this for us!

You can get AICc with the `AICc` function from *MuMIn*

```{r}
#| echo: true

# uncorrected AIC
AIC(mod_temp_precip)

# sample size corrected
AICc(mod_temp_precip)

```

::: {.fragment}

Our sample size is *large*, so the correction is small

:::


## Package *MuMIn* does other things 

*MuMIn* lets us do some things we maybe shouldn't

::: {.incremental}

- make all possible models with all possible variables
- then let AIC pick one, or average over all of them
- not recommended because we're just kind of "fishing" for something "significant" rather than testing hypotheses based in ecology/evolution/conservation
- *MuMIn* knows it's bad, their function for this workflow is called `dredge`

:::

## Package *MuMIn* does other things 

But `dredge`-ing can be useful for identifying which variables in our data are most "meaningful"


::: {.incremental}

- `dredge` up all models
- for each variable, sum up the weights of the models it is a part of
- normalize those summed weights to be [0, 1]
- the higher the summed weight, the more "important" that variable
- should do this based on variables of interest biologically, not just every variable you could possibly get your hands on

::::

## Package *MuMIn* does other things 

Variable importance via summed weights

```{r}
#| echo: true

# model with all possible variables
mod_big <- glm.nb(
    nspp ~ log(Plot_Area) + hii + avg_temp_annual_c + rain_annual_mm, 
    data = dat_plt, na.action = "na.fail"
)

# dredge it
big_dredge <- dredge(mod_big)

sw(big_dredge)
```

::: {.fragment}

So area is important (no surprise). But a new insight: human impact and rain are quite important, and temperature less so

:::



## References



