---
title: "Model Comparison and Averaging"

format:
  revealjs:
    slide-number: true
    preview-links: auto
    auto-stretch: false
    embed-resources: false
    controls: true
    html-math-method:
      method: mathjax
      url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-chtml-full.js"

bibliography: stat-mod.bib
csl: conservation-biology.csl

---

```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false

library(knitr)
library(ggplot2)
library(cowplot)
library(dplyr)
library(MASS)

opts_chunk$set(echo = FALSE)

# shared data setup used throughout
clim <- read.csv("data/plot_climate.csv")
geo_hii <- read.csv("data/hii_geo.csv")
tree <- read.csv("data/OpenNahele_Tree_Data.csv")

all_dat <- left_join(tree, clim, by = "PlotID") |>
    left_join(dplyr::select(geo_hii, !(lon:lat)), by = "PlotID")

dat_plt <- group_by(all_dat, PlotID) |>
    summarize(across(c(Island, Plot_Area, lon:elev_m), first),
              nspp = n_distinct(Scientific_name)) |>
    ungroup() |>
    filter(Plot_Area < 10000)
```


## What predicts species richness best?

***Temperature*** or ***Human Impact***?



::: {.fragment}

We saw that including both in one model had problem of collinearity

:::

::: {.fragment}

But does one predict richness better than another?

:::


::: {.fragment}

We cannot use LRT because these are not nested models, in fact they have the same d.f.

::: {.columns}

:::: {.column}

$$
\begin{aligned}
\log(\mu_\text{spp richness}) &= b_0 + b_1 \log(\text{area}) + b_2 \text{temperature} \\
\log(\mu_\text{spp richness}) &= b_0 + b_1 \log(\text{area}) + b_2 (\text{human impact})
\end{aligned}
$$

::::

:::

:::

## This is a job for AIC

***Akaike Information Criterion***


$$
AIC = -2 \mathcal{l}(\hat{\theta}_{MLE}) + 2k
$$

- $\mathcal{l}(\hat{\theta}_{MLE})$ is max of the log likelihood
- $k$ is number of parameters

::: {.fragment}

We can compare non-nested models with AIC

:::


## This is a job for AIC

$$
AIC = -2 \mathcal{l}(\hat{\theta}_{MLE}) + 2k
$$



```{r}
#| echo: true

mod_temp <- glm.nb(nspp ~ log(Plot_Area) + avg_temp_annual_c, 
                   data = dat_plt)
AIC(mod_temp)

mod_hii <- glm.nb(nspp ~ log(Plot_Area) + hii, 
                  data = dat_plt)
AIC(mod_hii)

```

The temperature model is better supported (lower AIC)


## This is a job for AIC

$$
AIC = -2 \mathcal{l}(\hat{\theta}_{MLE}) + 2k
$$

Purpose of $2k$ term is more clear for models with equal d.f.

```{r}
#| echo: true

mod_temp_precip <- glm.nb(
    nspp ~ log(Plot_Area) + avg_temp_annual_c + rain_annual_mm, 
    data = dat_plt
)
AIC(mod_temp_precip)

AIC(mod_hii)
```

## AIC penalizes more parameters

```{r}
#| echo: true

logLik(mod_temp_precip)
logLik(mod_temp)
```

Adding parameters will always increase log likelihood

::: {.fragment}

Is the increase real? Or just *"over fitting"*

:::


::: {.fragment}

*"over fitting"* is a term for what we know already: more parameters could let a model wiggle to fit noise

:::

## AIC penalizes more parameters

How? and why?


::: {.r-stack}

:::: {.fragment .fade-in-then-out}

Imagine you have one data set <span style="display:inline-block; width:0.8em; height:0.8em; background-color:black; border-radius:50%; vertical-align:middle;"></span> 

::: {.columns}

:::: {.column}

```{r}
#| fig-width: 6
#| fig-height: 6

set.seed(123)
x <- runif(40)
y <- 3 + x + rnorm(length(x), sd = 0.25)
g <- rep(c("old", "new"), each = 20)

dat <- data.frame(x, y, g)
gp <- ggplot(dat, aes(x, y, color = g)) +
    geom_point(size = 10) +
    geom_smooth(method = "lm", se = FALSE, 
                linewidth = 3, color = "black") +
    theme_cowplot() +
    theme(legend.position = "none", 
          axis.text = element_blank(), 
          axis.title = element_text(size = 34)) +
    scale_color_discrete(palette = c("transparent", "black"))

gp
```

::::

:::: {.column}

and fit a model to it by maximizing likelihood.

::::

:::

::::

:::: {.fragment .fade-in-then-out}

Imagine you have one data set <span style="display:inline-block; width:0.8em; height:0.8em; background-color:black; border-radius:50%; vertical-align:middle;"></span> then collect a new one <span style="display:inline-block; width:0.8em; height:0.8em; background-color:gray; border-radius:50%; vertical-align:middle;"></span>


::: {.columns}

:::: {.column}

```{r}
#| fig-width: 6
#| fig-height: 6

gp +
    geom_point(data = filter(dat, g == "new"), 
               aes(x, y), 
               color = "gray", size = 10)

```

::::

:::: {.column}

and calculate the log likelihood of the new data given the same model fit to the first data



::::

:::


::::

:::

## AIC penalizes more parameters

How? and why?


::: {.r-stack}

:::: {.fragment .fade-in-then-out}

Imagine you have one data set <span style="display:inline-block; width:0.8em; height:0.8em; background-color:black; border-radius:50%; vertical-align:middle;"></span> 

::: {.columns}

:::: {.column}

```{r}
#| fig-width: 6
#| fig-height: 6

set.seed(123)
x <- runif(40)
y <- 3 + x + rnorm(length(x), sd = 0.25)
g <- rep(c("old", "new"), each = 20)

dat <- data.frame(x, y, g)
gp <- ggplot(dat, aes(x, y, color = g)) +
    geom_point(size = 10) +
    geom_smooth(method = "lm", se = FALSE, 
                linewidth = 3, color = "black") +
    theme_cowplot() +
    theme(legend.position = "none", 
          axis.text = element_blank(), 
          axis.title = element_text(size = 34)) +
    scale_color_discrete(palette = c("transparent", "black"))

gp
```

::::

:::: {.column}

and fit a model to it by maximizing likelihood.

::::

:::

::::

:::

## AIC penalizes more parameters

How? and why?


Imagine you have one data set <span style="display:inline-block; width:0.8em; height:0.8em; background-color:black; border-radius:50%; vertical-align:middle;"></span> then collect a new one <span style="display:inline-block; width:0.8em; height:0.8em; background-color:gray; border-radius:50%; vertical-align:middle;"></span>


::: {.columns}

:::: {.column}

```{r}
#| fig-width: 6
#| fig-height: 6

gp +
    geom_point(data = filter(dat, g == "new"), 
               aes(x, y), 
               color = "gray", size = 10)

```

::::

:::: {.column}


::: {.r-stack}

:::: {.fragment .fade-in-then-out}

We could calculate the log likelihood for the new data ***but*** with parameter estimates from original model 

::::

:::: {.fragment}

The difference between the mean log likelihood of the original data and the mean log likelihood of the new data will be about $k$


::::


:::


::::

:::


:::

::: {.absolute top="70%" left="35%" style="z-index: 10;"}

::: {.fragment}

$$
\frac{1}{n} \mathcal{l}(\hat{\theta} \mid \text{data}_\text{orig}) - \frac{1}{n} \mathcal{l}(\hat{\theta} \mid \text{data}_\text{new}) \approx k
$$

:::

:::

## AIC penalizes more parameters

<br><br>

So AIC is trying to account for the chance that additional parameters are just fitting noise by including a penalty for those additional parameters


## Returning to model comparisons

We can rest assured that the improvement in AIC for the `temp + rain` model is not due just to adding antother parameter

```{r}
#| echo: true

AIC(mod_temp_precip)

AIC(mod_hii)
```

::: {.fragment}

A common rule is that models with AIC values within 2 points are effectively equivalent; $>2$ means one model really is better supported [@burnham-anderson]

:::


## Model averaging

::: {.incremental}

- model comparisons tells us which model is more supported
- model averaging is less about testing hypotheses and more about ***prediction***
- for example: how can we best predict tree richness across the pae ʻāina?

:::


## References


```{r}
#| eval: false
AIC(mod0, mod1)

sw(list(mod0, mod1))


fm1 <- lm(y ~ ., data = Cement, na.action = na.fail)
ms1 <- dredge(fm1)

# Sum of weights can be calculated/extracted from various objects:
sw(ms1)


```
