---
title: "Bayesian inference for complex models"

format:
  revealjs:
    slide-number: true
    preview-links: auto
    auto-stretch: false
    embed-resources: false
    controls: true
    html-math-method:
      method: mathjax
      url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-chtml-full.js"

bibliography: stat-mod.bib
csl: conservation-biology.csl

---

```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false

library(knitr)
library(ggplot2)
library(cowplot)
library(dplyr)
library(brms)
library(poilog)

opts_chunk$set(echo = FALSE)

# shared data setup used throughout
clim <- read.csv("data/plot_climate.csv")
geo_hii <- read.csv("data/hii_geo.csv")
tree <- read.csv("data/OpenNahele_Tree_Data.csv")

all_dat <- left_join(tree, clim, by = "PlotID") |>
    left_join(dplyr::select(geo_hii, !(lon:lat)), by = "PlotID") |> 
    mutate(Scientific_name = as.factor(Scientific_name))
```

```{r}
#| label: sad-make
#| cache: true
#| include: false

dat_sad <- group_by(all_dat, PlotID, Scientific_name, .drop = FALSE) |>
    summarize(across(c(Island, Plot_Area, lon:elev_m), first),
              abund = n()) |>
    ungroup() |>
    filter(Plot_Area < 10000)
```

## The species abundnace distribution

Ecology has spent a long time thinking about the probability distribution of abundances

The shape of the SAD 

::: {incremental}

- figured in early and influential debates about ecological processes [@fisher1943; @preston1962]
- is the focus of the unified neutral theory of biodiversity [@hubbell2011]
- is still central to models of biodiversity, though we now know other biodiversity dimensions are important [@overcast2021]

:::

## The species abundnace distribution

Ecology has spent a long time thinking about the probability distribution of abundances

::: {.columns}

:::: {.column}

::: {.fragment}

Most species are rare

<br>

```{r}
#| fig-width: 5
#| fig-height: 4

x <- seq(log(1), log(1000), length.out = 30) |> 
    exp() |> 
    round()
dat_sing <- data.frame(x, px = dpoilog(x, 10, 4))

ggplot(dat_sing, aes(x, px)) +
    geom_point(size = 2) +
    scale_x_log10() +
    xlab("Abundance (note log scale)") +
    ylab("Probability") +
    theme(axis.text = element_text(size = 18), 
          axis.title = element_text(size = 22))
```

:::

::::

:::: {.column}

::: {.fragment}

But perhaps singletons are not the most frequent

```{r}
#| fig-width: 5
#| fig-height: 4

dat_hump <- data.frame(x, px = dpoilog(x, 3, 1.2))

ggplot(dat_hump, aes(x, px)) +
    geom_point(size = 2) +
    scale_x_log10() +
    xlab("Abundance (note log scale)") +
    ylab("Probability") +
    theme(axis.text = element_text(size = 18), 
          axis.title = element_text(size = 22))
```

:::

::::

:::


## The species abundnace distribution

Ecology has spent a long time thinking about the probability distribution of abundances

::: {.columns}

:::: {.column}


Log series distribution

<br>

```{r}
#| fig-width: 5
#| fig-height: 4

x <- seq(log(1), log(1000), length.out = 30) |> 
    exp() |> 
    round()
dat_sing <- data.frame(x, px = dpoilog(x, 10, 4))

ggplot(dat_sing, aes(x, px)) +
    geom_point(size = 2) +
    scale_x_log10() +
    xlab("Abundance (note log scale)") +
    ylab("Probability") +
    theme(axis.text = element_text(size = 18), 
          axis.title = element_text(size = 22))
```


::::

:::: {.column}

Log normal distribution

<br>

```{r}
#| fig-width: 5
#| fig-height: 4

dat_hump <- data.frame(x, px = dpoilog(x, 3, 1.2))

ggplot(dat_hump, aes(x, px)) +
    geom_point(size = 2) +
    scale_x_log10() +
    xlab("Abundance (note log scale)") +
    ylab("Probability") +
    theme(axis.text = element_text(size = 18), 
          axis.title = element_text(size = 22))
```


::::

:::


## The species abundnace distribution

Ecology has spent a long time thinking about the probability distribution of abundances

Turns out the ***Poisson log normal*** can accamodate both shapes and a gradient in between

::: {.columns}

:::: {.column}



```{r}
#| fig-width: 5
#| fig-height: 4

x <- seq(log(1), log(1000), length.out = 30) |> 
    exp() |> 
    round()
dat_sing <- data.frame(x, px = dpoilog(x, 10, 4))

ggplot(dat_sing, aes(x, px)) +
    geom_point(size = 2) +
    scale_x_log10() +
    xlab("Abundance (note log scale)") +
    ylab("Probability") +
    theme(axis.text = element_text(size = 18), 
          axis.title = element_text(size = 22))
```


::::

:::: {.column}


```{r}
#| fig-width: 5
#| fig-height: 4

dat_hump <- data.frame(x, px = dpoilog(x, 3, 1.2))

ggplot(dat_hump, aes(x, px)) +
    geom_point(size = 2) +
    scale_x_log10() +
    xlab("Abundance (note log scale)") +
    ylab("Probability") +
    theme(axis.text = element_text(size = 18), 
          axis.title = element_text(size = 22))
```


::::

:::


## The species abundnace distribution

The ***Poisson log normal*** 

$$
\begin{aligned}
\log(\lambda_i) &\sim \mathscr{N}(\mu, ~\sigma) \\
y_i &= \text{Pois}(\lambda_i)
\end{aligned}
$$

::: {.fragment}

Some familiar looking things

:::: {.incremental}

- log link function for mean of Poisson 
- but the mean comes from a distribution like a random effect
- adding some explanatory variables will reveal the random effect structure

::::


:::

## The species abundnace distribution

The ***Poisson log normal*** as a random effects model

$$
\begin{aligned}
\mu_i &= b_0 + b_1 x_{1i} \\
\log(\lambda_i) &\sim \mathscr{N}(\mu_i, ~\sigma) \\
y_i &= \text{Pois}(\lambda_i)
\end{aligned}
$$

::: {.fragment}

Due to additive property of normal we can rearrange

$$
\begin{aligned}
b_{\text{sp} ~i} &\sim \mathscr{N}(0, ~\sigma_\text{sp}) \\
\log(\lambda_i) &= b_0 + b_1 x_1 + b_{\text{sp} ~i} \\
y_i &= \text{Pois}(\lambda_i)
\end{aligned}
$$

:::




## The species abundnace distribution

The ***Poisson log normal*** as a random effects model



$$
\begin{aligned}
b_{\text{sp} ~i} &\sim \mathscr{N}(0, ~\sigma_\text{sp}) \\
\log(\lambda_i) &= b_0 + b_1 x_1 + b_{\text{sp} ~i} \\
y_i &= \text{Pois}(\lambda_i)
\end{aligned}
$$



This is a ***random intercepts*** model, each species gets a random intercept


## The species abundnace distribution

The ***Poisson log normal*** as a random effects model

::: {.columns}

:::: {.column}

At one sampling location, distribution of abundances will be distributed Poisson log normal

::: {.r-stack}

::: {.fragment .fade-in-then-out}

You might be thinking about overdispersion, *Poisson log normal* is already "overdispersed" and, in fact, quite similar to negative binomial

:::

::: {.fragment}

But what about SAD at another sampling location??

:::



:::

::::

:::: {.column}

```{r}
#| fig-width: 5
#| fig-height: 3

ggplot(dat_hump, aes(x, px)) +
    geom_point(size = 2) +
    scale_x_log10() +
    xlab("Abundance (note log scale)") +
    ylab("Probability") +
    theme(axis.text = element_text(size = 18), 
          axis.title = element_text(size = 22))

```

::: {.fragment style="margin-top: -2em;"}

```{r}
#| fig-width: 5
#| fig-height: 3

ggplot(data.frame(x, px = dpoilog(x, 1, 3)), aes(x, px)) +
    geom_point(size = 2) +
    scale_x_log10() +
    xlab("Abundance (note log scale)") +
    ylab("Probability") +
    theme(axis.text = element_text(size = 18), 
          axis.title = element_text(size = 22))

```

:::

::::

:::

## The species abundnace distribution

The ***Poisson log normal*** as a random effects model



$$
\begin{aligned}
b_{\text{sp} ~i} &\sim \mathscr{N}(0, ~\sigma_\text{sp}) \\
b_{\text{site} ~j} &\sim \mathscr{N}(0, ~\sigma_\text{site}) \\
\log(\lambda_{ij}) &= b_0 + b_1 x_1 + b_{\text{sp} ~i} + b_{\text{site} ~j} \\
y_{ij} &= \text{Pois}(\lambda_{ij})
\end{aligned}
$$



We need ***random intercepts*** for species and site


::: {.fragment}

This model means that a common species will tend to be common everywhere (and a rare species rare everywhere), but there is random variation site-to-site due to demographic noise

:::


## The species abundnace distribution

The ***Poisson log normal*** as a random effects model

::: {.fragment}

Why this is a challenging model

::: {.incremental}

- Random effects models are already a challenge
    - There are `r n_distinct(dat_sad$PlotID)` plots
    - There are `r n_distinct(dat_sad$Scientific_name)` species
    - Estimating (via likelihood or posterior) $\sigma_\text{sp}$ and $\sigma_\text{site}$ requires integrating over all those entities
- Mixing Poisson and (log) normal makes the necessary integrals all the more challenging
- Using MCMC means we do not need to do these integrals

:::

:::



## Fitting the SAD with `brm`

Here is the code for ***just the random effects***

```{r}
#| label: fit-brms
#| cache: true
#| echo: true


sad_mod <- brm(
    abund ~ (1 | Scientific_name) +
        (1 | PlotID), 
    data = dat_sad,
    family = poisson(),
    prior = c(
        prior(normal(0, 100), class = Intercept),
        prior(exponential(0.01), class = sd)
    ),
    chains = 4,
    cores = 4,
    seed = 123, 
    iter = 4000, 
    warmup = 1000
)
```

## Fitting the SAD with `brm`

How does it look?

```{r}
summary(sad_mod)


```


## Modeling species abundances

With this approach we can now ask more interesting questions

::: {.incremental}

- are native or non-native taxa more abundant?
- what variables determine the balance of native and non-native taxa?


:::

## References



