---
format:
  html:
    code-links:
      - text: View source
        icon: github
        href: https://github.com/uhm-biostats/stat-mod/blob/main/binom-glm.qmd
---


# The Binomial Generalized Linear Model


```{r}
#| label: setup
#| include: false

library(knitr)
library(dplyr)
library(ggplot2)
library(cowplot)
library(ggpointdensity)
```

So far we have worked with linear models that assume a normal distribution. Whether or not we consider random effects, the model is built off the idea that the mean of a normal distribution changes across values of the explanatory variables and from that normal distribution we get our response data. In the simple linear model case that approach looks like this:

In graphical form

```{r}
set.seed(123)
x <- runif(40)
y <- rnorm(length(x), x, sd = 0.25)

ggplot(data.frame(x, y), aes(x, y)) +
    geom_point() +
    geom_smooth(method = "lm", se = FALSE) +
    theme_cowplot() +
    theme(axis.text = element_blank(), 
          axis.title = element_text(size = 20)) 
```

And in equation form

$$
\begin{aligned}
\bar{y} &= b_0 + b_1 x \\
y_i &\sim \mathcal{N}(\bar{y},~\sigma)
\end{aligned}
$$

$\bar{y}$ is the blue line in the figure, and each data point is presumed to come from a normal distribution with mean $\bar{y}$. But you may wonder, does the distribution always have to be a normal? The answer is no! Many other distributions can be used for these kinds of linear models. When we use a distribution other than normal, we call the model a *generalized* linear model (GLM). To introduce how GLMs work we'll start with the binomial GLM, also known as logistic regression.

## Binomial generalized linear models




## Slides {#sec-slides}


```{=html}
<iframe class="slide-deck" src="binom-glm-slides.html"></iframe>
```
