---
title: "The Binomial Generalized Linear Model"
---

```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false

library(knitr)
library(dplyr)
library(ggplot2)
library(cowplot)
library(ggpointdensity)

opts_chunk$set(echo = FALSE)
```


## Recap of normal linear models

```{r}
lmfig <- function(ngrp, ri, rs) {
    ntot <- 60
    n <- ntot / ngrp + 2
    
    x <- replicate(
        ngrp, {
            x0 <- runif(n - 2)
            c(0, x0, 1)
        }, 
        simplify = FALSE
    ) |> 
        unlist()
    
    yhat <- 1 + 1.5 * x
    
    if(ri | rs) {
        g <- (seq(0, ngrp, length.out = ngrp) / ngrp * 1.5) |> 
            rep(each = n)
        yhat <- yhat + g
    } else {
        g <- rep(0, length(yhat))
    }
    
    if(rs) {
        yhat <- yhat + x * (rev(0.75 * g) - 1)
    }

    y <- yhat + rnorm(length(yhat), sd = 0.25)
    
    dat <- data.frame(x, y, yhat, g = as.factor(g))
    
    ggplot(dat, aes(x, y, color = g)) +
        geom_point() +
        scale_color_viridis_d() +
        geom_line(mapping = aes(x, yhat)) +
        theme_cowplot() +
        theme(legend.position = "none", 
              axis.text = element_blank(), 
              axis.title = element_text(size = 20)) 
}

```



::: {.columns}

:::: {.column width=33%}

**Simple linear model**

```{r}
#| fig-width: 3
#| fig-height: 3

lmfig(1, FALSE, FALSE)
```

No random effects

::::

:::: {.column width=33%}

**Linear model with random intercepts**

```{r}
#| fig-width: 3
#| fig-height: 3

lmfig(3, TRUE, FALSE)
```

Groups (usually more than 3) affect intercepts

::::

:::: {.column width=33%}

**Linear model with random slopes**

```{r}
#| fig-width: 3
#| fig-height: 3

lmfig(3, TRUE, TRUE)
```

Groups (usually more than 3) affect intercepts and slopes

::::

:::

## Recap of normal linear models



**Simple linear model**

$$
\begin{aligned}
\bar{y} &= b_0 + b_1 x \\
y_i &\sim \mathcal{N}(\bar{y},~\sigma)
\end{aligned}
$$

## Recap of normal linear models

**Linear model with random intercepts**

$$
\begin{aligned}
r_{0}^{(group~k)} &\sim \mathcal{N}(0,~\sigma_{r_0}) \\
\bar{y} &= b_0 + r_0^{(group~k)} + b_1 x \\
y &\sim \mathcal{N}(\bar{y},~\sigma)
\end{aligned}
$$


## Recap of normal linear models

**Linear model with random slopes**

$$
\begin{aligned}
r_0^{(group~k)} &\sim \mathcal{N}(0,~\sigma_{r_0}) \\
r_1^{(group~k)} &\sim \mathcal{N}(0,~\sigma_{r_1}) \\
\bar{y} &= b_0 + r_0^{(group~k)} + (b_1 + r_1^{(group~k)}) x \\
y &\sim \mathcal{N}(\bar{y},~\sigma)
\end{aligned}
$$

(technically we usually model a correlation term between $r_0$ and $r_1$)




## Recap of normal linear models 

<br><br>

But what's so special about normal distribution?

::: {.fragment}

...nothing, we can use other distributions

:::


## Introducing: generalized linear models

<br><br>

When we use a distribution other than a normal we call our model a ***generalized linear model***

::: {.fragment}

We'll start with the binomial as an example 

:::


## Binomial generalized linear models

These data are not normal, and fitting a straight line doesn't make sense

```{r}
#| fig-width: 6
#| fig-height: 5
#| fig-align: center

set.seed(123)
x <- runif(40)
y <- rbinom(length(x), size = 1, prob = plogis(10 * (x - 0.5)))

dat <- data.frame(x, y)

ggplot(dat, aes(x, y)) +
    geom_point(size = 4) +
    geom_smooth(method = "lm", se = FALSE) +
    theme_cowplot() +
    theme(axis.text = element_blank(), 
          axis.title = element_text(size = 30))
```


## Binomial generalized linear models

Recognizing a binomial distribution and a model bounded between [0, 1] makes more sense

```{r}
#| fig-width: 6
#| fig-height: 5
#| fig-align: center

gpbiom <- ggplot(dat, aes(x, y)) +
    geom_point(size = 4) +
    geom_smooth(method = "glm", se = FALSE, 
                method.args = list(family = binomial)) +
    theme_cowplot() +
    theme(axis.text = element_blank(), 
          axis.title = element_text(size = 30))

gpbiom
```

## Binomial generalized linear models


We've been here before: modeling presence/absence across a gradient using "expit" and "logit" functions

$$
p = \frac{1}{1 + e^{-(b_0 + b_1 x)}}
$$
In GLM this kind of equation is called the ***link function***

::: {.fragment}
And we can use $p$ in a binomial distribution

$$
y \sim \text{Binom}(p,~N = 1)
$$

:::




## Understanding the link function

::: {.columns}

:::: {.column}

<br>

This is called the *expit* function

$$
p = \frac{1}{1 + e^{-(b_0 + b_1 x)}}
$$

and produces the curve we just saw

::::

:::: {.column}

<br>

```{r}
#| fig-width: 6
#| fig-height: 5
#| fig-align: center

ggplot(data.frame(x = 0:1, p = 0:1), aes(x, p)) +
    geom_function(fun = function(x) plogis(10 * (x - 0.5)), 
                  linewidth = 2) +
    theme_cowplot() +
    theme(axis.text = element_blank(), 
          axis.title = element_text(size = 30))
```

::::

:::

## Understanding the link function

::: {.columns}

:::: {.column}


Its often easier to think in terms of the *logit* function

$$
\text{logit}(p) = b_0 + b_1 x
$$

which is the inverse of the *expit* function

::::

:::: {.column}


```{r}
#| fig-width: 6
#| fig-height: 5
#| fig-align: center

ggplot(data.frame(x = 0:1, p = 0:1), aes(x, p)) +
    geom_function(fun = function(x) (1 * (x - 0.5)), 
                  linewidth = 2) +
    theme_cowplot() +
    ylab("logit(p)") +
    theme(axis.text = element_blank(), 
          axis.title = element_text(size = 30))
```

::::

:::

Now we just have a straight line! The link function gives us the linear in a generalized ***linear*** model

## Generalized linear models in R: `glm`

<br>

How do we fit GLMs in R, why with the `glm` function!

```{r}
#| echo: true

mod <- glm(y ~ x, data = dat, family = binomial)

```

::: {.fragment}

Compared to `lm` there is one new thing: `family = binomial`

:::

::: {.fragment}

The argument `family = binomial` tells `glm` two things:

1. we're assuming a binomial distribution
2. we want to use a *logit* link function (though other link functions are available)

:::



## Interpreting coefficients

<br>

::: {.columns}

:::: {.column}

We can see the coefficients like this

```{r}
#| echo: true

coefficients(mod)

```

But what do they mean?


::::

:::: {.column}

::: {.fragment}

```{r}
#| fig-width: 5
#| fig-height: 4

gpbiom +
    theme(axis.text = element_text(size = 24))
```

How could intercept be `r round(coefficients(mod)[1], 3)`?

:::

::::

:::


## Interpreting coefficients

<br>

::: {.columns}

:::: {.column}

We can see the coefficients like this

```{r}
#| echo: true

coefficients(mod)

```

But what do they mean?


::::

:::: {.column}


```{r}
#| fig-width: 5
#| fig-height: 4

gplink <- ggplot(cbind(dat, p = predict(mod)), aes(x, p)) +
    geom_line(linewidth = 2) +
    ylab("logit(p)") +
    theme_cowplot() +
    theme(axis.title = element_text(size = 30), 
          axis.text = element_text(size = 24))

gplink
```

Coefficients make sense on link function scale


::::

:::



## Visualizing the model and data

*ggplot2* is ready to help


```{r}
#| echo: true
#| eval: false

ggplot(dat, aes(x, y)) +
    geom_point() +
    geom_smooth(method = "glm",
                method.args = list(family = binomial))

```

::: {.fragment}

*(with some extra modification)*

```{r}
#| fig-width: 6
#| fig-height: 4

ggplot(dat, aes(x, y)) +
    geom_point(size = 4) +
    geom_smooth(method = "glm", 
                method.args = list(family = binomial)) + 
    theme_cowplot() +
    theme(axis.title = element_text(size = 30), 
          point = element_point(size = 8))

```

:::


## Visualizing the model and data


::: {.columns}

:::: {.column width=40%}

Common issue: points can be hard to see

- Jitter can help


::::

:::: {.column width=60%}



```{r}
#| echo: true
#| eval: false

# only in vertical direction
ggplot(dat, aes(x, y)) +
    geom_jitter(width = 0, height = 0.05) 

```

<br>

```{r}
#| fig-width: 6
#| fig-height: 4

ggplot(dat, aes(x, y)) +
    geom_jitter(width = 0, height = 0.05, size = 4) +
    theme_cowplot() +
    theme(axis.title = element_text(size = 30), 
          point = element_point(size = 8))

```



::::

:::



## Visualizing the model and data


::: {.columns}

:::: {.column width=40%}

Common issue: points can be hard to see

- Jitter can help
- Transparency is an option



::::

:::: {.column width=60%}



```{r}
#| echo: true
#| eval: false

ggplot(dat, aes(x, y)) +
    geom_point(alpha = 0.5) 

```

<br>

```{r}
#| fig-width: 6
#| fig-height: 4

ggplot(dat, aes(x, y)) +
    geom_point(alpha = 0.5, size = 4) +
    theme_cowplot() +
    theme(axis.title = element_text(size = 30), 
          point = element_point(size = 8))

```



::::

:::


## Visualizing the model and data


::: {.columns}

:::: {.column width=40%}

Common issue: points can be hard to see

- Jitter can help
- Transparency is an option
- Visualizing local density is pretty


::::

:::: {.column width=60%}



```{r}
#| echo: true
#| eval: false

library(ggpointdensity)
ggplot(dat, aes(x, y)) +
    geom_pointdensity(method = "kde2d") +
    scale_color_viridis_c()

```

<br>

```{r}
#| fig-width: 6
#| fig-height: 4


ggplot(dat, aes(x, y)) +
    geom_pointdensity(method = "kde2d", size = 4) +
    theme_cowplot() +
    scale_color_viridis_c() + 
    theme(axis.title = element_text(size = 30), 
          point = element_point(size = 8))

```



::::

:::


## Visualizing the model and data

Let's look at the link function in two ways

1. With the expit equation 

```{r}
#| echo: true
#| eval: false

# get link function intercept and slope
bb <- coefficients(mod)
bb

# write a function for the equation
my_expit <- function(x, intercept, slope) {
    intercept + slope * x
}

# use `geom_function`
ggplot(data.frame(x = 0:1, y = 0:1), aes(x, y)) +
    geom_function(fun = my_expit, 
                  args = list(intercept = bb[1], 
                              slope = bb[2])) +
    ylab("logit(p)")
```

## Visualizing the model and data

Let's look at the link function in two ways

2. With the predicted values for *logit(p)*

```{r}
#| echo: true
#| eval: false

# function `predict` gives us model predictions
# `type = link` says, give us predictions on link function scale
logitp_hat <- predict(mod, type = "link")
head(logitp_hat)

# add a column for logit(p) and you're ready to plot!
ggplot(mutate(dat, logitp = logitp_hat), 
       aes(x, logitp)) +
    geom_line(linewidth = 2) +
    ylab("logit(p)")



```

## Visualizing the model and data

Either way gives you this graph

```{r}
#| fig-width: 8
#| fig-height: 6
#| fig-align: center

gplink

```

## Binomial GLM: Use cases

```{r}
#| label: data-prep

tree <- read.csv("data/OpenNahele_Tree_Data.csv")
geo_hii <- read.csv("data/hii_geo.csv")

tree_dat <- left_join(tree, geo_hii)

tree_stem <- select(tree_dat, PlotID, elev_m, Native_Status) |> 
    mutate(non_nat = 1 * (Native_Status == "alien")) |> 
    select(!Native_Status)

tree_plot <- group_by(tree_dat, PlotID) |> 
    summarize(elev_m = first(elev_m), 
              any_non = any(Native_Status == "alien") * 1, 
              num_non = sum(Native_Status == "alien"), 
              num_tre = n()) |> 
    ungroup() |> 
    as.data.frame()

stra_gua <- group_by(tree_dat, PlotID) |> 
    summarize(elev_m = first(elev_m), 
              any_gua = any(Scientific_name == "Psidium cattleianum") * 1, 
              num_gua = sum(Scientific_name == "Psidium cattleianum"), 
              num_tre = n()) |> 
    ungroup() |> 
    as.data.frame()
```


::: {.fragment}

**presence/absence**, e.g., where is the strawberry guava?

:::


::: {.columns}

:::: {.column}

::::: {.fragment}

```{r}
#| echo: true

# i did some data prep already
head(stra_gua) |> kable()


```

:::::

::::

:::: {.column}

::::: {.fragment}

```{r}
#| echo: true

mg <- glm(any_gua ~ elev_m, 
          data = stra_gua, 
          family = binomial)

coefficients(mg)
```

```{r}
#| fig-width: 8
#| fig-height: 6

ggplot(stra_gua, aes(elev_m, any_gua)) +
    geom_pointdensity(size = 4, 
                      position = position_jitter(width = 0, height = 0.05)) +
    scale_color_viridis_c() +
    geom_smooth(method = "glm", color = "black",
                method.args = list(family = binomial)) +
    xlab("Elevation (m)") +
    ylab("Presence = 1") +
    theme_cowplot() +
    theme(axis.text = element_text(size = 24), 
          axis.title = element_text(size = 30), 
          legend.text = element_text(size = 24),
          legend.title = element_text(size = 24),
          legend.key.height = unit(3, "lines"), 
          legend.key.width = unit(2, "lines"))
```

:::::

::::

:::



## Binomial GLM: Use cases



**binary property**, e.g., non-native or native?



::: {.columns}

:::: {.column}

::::: {.fragment}

```{r}
#| echo: true

# i did some data prep already
head(tree_plot) |> kable()


```

:::::

::::

:::: {.column}

::::: {.fragment}

```{r}
#| echo: true

mn <- glm(any_non ~ elev_m, 
          data = tree_plot, 
          family = binomial)

coefficients(mn)
```

```{r}
#| fig-width: 8
#| fig-height: 6

ggplot(tree_plot, aes(elev_m, any_non)) +
    geom_pointdensity(size = 4, 
                      position = position_jitter(width = 0, height = 0.05)) +
    scale_color_viridis_c() +
    geom_smooth(method = "glm", color = "black",
                method.args = list(family = binomial)) +
    xlab("Elevation (m)") +
    ylab("Presence of non-native") +
    theme_cowplot() +
    theme(axis.text = element_text(size = 24), 
          axis.title = element_text(size = 30), 
          legend.text = element_text(size = 24),
          legend.title = element_text(size = 24),
          legend.key.height = unit(3, "lines"), 
          legend.key.width = unit(2, "lines"))
```

:::::

::::

:::


## Binomial GLM: Use cases



**binary property**, wait, aren't we wasting tree-level data?



::: {.columns}

:::: {.column}



```{r}
#| echo: true

# i did some data prep already
head(tree_stem) |> kable()


```



::::

:::: {.column}

::::: {.fragment}

```{r}
#| echo: true

ms <- glm(non_nat ~ elev_m, 
          data = tree_stem, 
          family = binomial)

coefficients(ms)
```

```{r}
#| fig-width: 8
#| fig-height: 6

ggplot(tree_stem, aes(elev_m, non_nat)) +
    geom_pointdensity(size = 4, 
                      position = position_jitter(width = 0, height = 0.05)) +
    scale_color_viridis_c() +
    geom_smooth(method = "glm", color = "black",
                method.args = list(family = binomial)) +
    xlab("Elevation (m)") +
    ylab("Presence of non-native") +
    theme_cowplot() +
    theme(axis.text = element_text(size = 24), 
          axis.title = element_text(size = 30), 
          legend.text = element_text(size = 24),
          legend.title = element_text(size = 24),
          legend.key.height = unit(3, "lines"), 
          legend.key.width = unit(2, "lines"))
```

:::::

::::

:::


## Binomial GLM: Use cases

**binary property**, e.g., non-native or native?

<br>

No that wasn't a good idea---pseudo replication in plots

<br>

::: {.fragment}
But before you go thinking of random effects

:::

<br>

::: {.fragment}
Remember, within each plot is like a binomial sample:

- "successes": number of non-native trees
- total size: number of all trees

:::


## Binomial GLM: Use cases



**number of successes vs. failures**, back to plot-level data



::: {.columns}

:::: {.column}



```{r}
#| echo: true
#| eval: false

head(tree_plot) |> kable()
# hint: scroll right

```

```{r}
head(tree_stem) |> kable()
```



::::

:::: {.column}

::::: {.fragment}

```{r}
#| echo: true

# wants a matrix (from cbind) of num success and num failure
mns <- glm(cbind(num_non, num_tre - num_non) ~ elev_m, 
          data = tree_plot, 
          family = binomial)

coefficients(mns)
```

:::::

::::: {.fragment}

Plotting is more convoluted so let's go into it more

:::::



::::

:::


## Binomial GLM: Use cases

**number of successes vs. failures**, back to plot-level data

```{r}
#| echo: true
#| eval: false

ggplot(tree_plot, aes(elev_m, num_non / num_tre, 
                      # include `success` and `failure`
                      # for use in `geom_smooth`
                      success = num_non, 
                      failure = num_tre - num_non)) +
    geom_pointdensity() +
    scale_color_viridis_c() +
    geom_smooth(method = "glm", color = "black",
                # need to add a specific formula
                formula = cbind(success, failure) ~ x,
                method.args = list(family = binomial))
```


## Binomial GLM: Use cases

**number of successes vs. failures**, back to plot-level data

*(with some extra beautifying)*

```{r}
#| fig-width: 8
#| fig-height: 6
#| fig-align: center

ggplot(tree_plot, aes(elev_m, num_non / num_tre, 
                      success = num_non, 
                      failure = num_tre - num_non)) +
    geom_pointdensity(size = 4) +
    scale_color_viridis_c() +
    geom_smooth(method = "glm", color = "black",
                formula = cbind(success, failure) ~ x,
                method.args = list(family = binomial)) +
    xlab("Elevation (m)") +
    ylab("Proportion non-native") +
    theme_cowplot() +
    theme(axis.text = element_text(size = 24), 
          axis.title = element_text(size = 30), 
          legend.text = element_text(size = 24),
          legend.title = element_text(size = 24),
          legend.key.size = unit(3, "lines"))
```







