---
format:
  html:
    code-links:
      - text: View source
        icon: github
        href: https://github.com/uhm-biostats/stat-mod/blob/main/bringing-it-together.qmd
---

# CARE and FAIR in presenting data with dynamic documents

This lab will work through planning and envisioning CARE and FAIR approaches to data in Hawaiʻi as well as visualizing and numerically summarizing the data we will use in this course. You can download the template `.qmd` file for this lab here:

{{< downloadthis labs/lab01-data.qmd dname=lab01-data label="Download lab template">}}

```{r}
#| include: false

# the lab qmd file
f <- "labs/lab01-data.qmd"
l <- readLines(f)

# the yaml
y <- grep("---", l) |> 
    (\(x) x[1]:x[2])() |> 
    (\(x) l[x])()

# only the body
b <- l[!(l %in% y)] |> 
    gsub("\\*Delete this .* once complete\\*", "", x = _)

# write only the body for use with `{{< include %s >}}`
# fb <- gsub("\\.qmd", "_bonly.qmd", f)
fb <- tempfile(fileext = ".qmd")

writeLines(b, fb)
```

Note: the yaml header of this file is a little more complex than we have seen before


```{r}
#| echo: false
#| classes: s-output

cat(y, sep = "\n")
```

You now can see how to cleanly include multiple authors. We also see our first example of customizing PDF output with `\usepackage` which is how we load a LaTeX package (same idea as an R package), in this case to support directly typing the unicode ʻokina.

The template will prompt you to work through the following sections.


```{r}
#| echo: false
#| results: asis

m <- knitr::knit_child(fb, options = list(include = FALSE), 
                       quiet = TRUE)

cat(m, sep = "\n")
```

