# CARE and FAIR in presenting data with dynamic documents on GitHub


## CARE and FAIR consideration in Hawaiʻi

Write out your manaʻo on CARE and FAIR in Hawaiʻi using class materials, class readings, and most especially your previously assigned written reflections on those readings.

### Ideas for making data from Hawaiʻi CARE and FAIR



## Describing *OpenNahele* and *Hawaiʻi Climate Data Portal Data*

Delete this list once complete

- read-in and join all data like we did in the R refresher lab (fine to copy paste over) for subsequent use
- create a plot-level data object like we did in the R refresher lab (fine to copy paste over) for subsequent use
- across the entire data set (*not* broken up by island or anything else) calculate 25%, 50%, and 75% quantiles for each numeric plot-level variable but not `Year` or `Census` or `lon` or `lat`. *Hints:* you can use the `quantile` function, you will need to set `na.rm = TRUE`, you can make a `data.frame` by hand if that makes most sense, or you can use `across` in combination with `reframe` (which is the *dplyr* approach)
- use a ridgeline plot or stacked histograms to show distributions for all species richness, basal area, and abundance variables by island



```{r}
#| eval: false
#| echo: false


all_dat_tib |> 
    reframe(across(Island:elev_m & where(is.numeric) & !(Year:lat), quantile, probs = c(0.25, 0.5, 0.75), na.rm = TRUE))
```



## Relating tree diversity to environmental predictors

Delete this list once complete

- create new plot-level columns for species richness---not proportions---that corrects for area (should be, e.g., `log(nspp) / log(Plot_Area)`)
- make a figure for the response of species richness to plot area (as we did in R refresher lab, fine to copy paste)
- make a figure showing, across all islands, proportion of plots with $\geq 75\%$ native trees (as we did in R refresher lab, fine to copy paste)
- choose three other figures to make where you plot any kind of tree-related response variable (e.g. species richness) against an environmental predictor variable


```{r}
#| eval: false
#| echo: false


dat_by_plot <- mutate(dat_by_plot, 
                      log_nspp_log_area = log(nspp) / log(Plot_Area)) 

filter(dat_by_plot, Plot_Area < 10000) |> 
    ggplot(aes(Plot_Area, log_nspp_log_area)) +
    geom_hex() +
    scale_x_continuous(transform = "log10")
```


