# Data we will work with

Jump to [slides](#sec-slides).

## Intorducing the data

We will be working with data from forest monitoring plots across the pae ʻāina Hawaiʻi as well as environmental and geophysical data measured or interpolated at these monitoring plots. The forest monitoring plot data are organized and distributed in a data resource called [*OpenNahele*](https://datadryad.org/dataset/doi:10.5061/dryad.1kk02qr){target="_blank"} [@craven2018]. 

In this context "open" refers to the data being openly shared and accessible in alignment with the FAIR principles which we discuss in the [next section](care-fair.qmd). [Nahele](https://wehe.hilo.hawaii.edu/?q=nahele#w2w2-15258){target="_blank"} in ʻōlelo Hawaiʻi means, in this context, forest. The use of ʻōlelo Hawaiʻi points to the fact that the data pertain to the [pae ʻāina](https://wehe.hilo.hawaii.edu/?q=pae%20aina#w2w2-18645) Hawaiʻi. This naming choice could show respect and acknowledgement of the Hawaiian provenance of these data, but could also be seen as appropriation. The interpretation of the naming choice depends in part on whether the data not only comply with FAIR, but also the CARE principles, which, again, we discuss in the [next section](care-fair.qmd). The CARE principles mandate that Indigenous provenance and governance rights are acknowledged and respected in the stewardship of Indigenous data. The data in *OpenNahele* originate from the pae ʻāina Hawaiʻi and thus within the traditional lands of the [Kanaka](https://wehe.hilo.hawaii.edu/?q=kanaka#w2w2-5691) [ʻŌiwi](https://wehe.hilo.hawaii.edu/?q=oiwi#w2w2-29101), making them Indigenous data. Without attribution, or any indication about ʻŌiwi protocols and rights for data collection and use, *OpenNahele* does not meet the CARE principles. While the good intentions of those distributing the data is not being questioned, we also recognize there is opportunity for making right the stewardship of these, and other, data from Hawaiʻi.  We dive much deeper into this discussion in the next section on [CARE and FAIR](care-fair.qmd).

The environmental and geophysical data we will use include climate variables, information about human impact, elevation, and geologic age of substrates at the locations of the plot data. For climate variables we use the [Hawaiʻi Climate Data Portal](https://www.hawaii.edu/climate-data-portal/climate-atlas/){target="_blank"} [@mclean2023]. Elevation, geologic age, and human impact data have been organized by the authors of the *OpenNahele* data set in a separate data publication [@craven2019] and we will use those already compiled data rather than re-gather them from the primary sources. However, for the sake of completeness, the primary sources are as follows: elevation [@jarvis2008], substrate age [@sherrod2007], and human impact [@sedac2005].

While these environmental and geophysical data meet the FAIR principles, they again are not CARE compliant. Part of our work in this course will be discussing and envisioning how data collected in Hawaiʻi can live up to the CARE principles.

## Preliminary description of the data

We won't go deep into visualizing and numerically describing the data here because we'll save that for our [R refresher](r-refresh.qmd) when we'll review coding tools by working with these data.

```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false


library(sf)
library(ggplot2)
library(knitr)
library(dplyr)

opts_chunk$set(echo = FALSE, 
                      message = FALSE, 
                      warning = FALSE)
```

```{r}
#| label: data-setup

load("data/hi.rda")
tree <- read.csv("data/OpenNahele_Tree_Data.csv")
clim <- read.csv("data/plot_climate.csv")
clim_st <- st_as_sf(clim, coords = c("lon", "lat"), 
                    crs = st_crs(hi))
hii_geo <- read.csv("data/hii_geo.csv")

```


### Forest plot data

The forest plot data are found in `data/OpenNahele_Tree_Data.csv`. The data contain taxonomic identity and diameter at breast height (DBH) for `r nrow(tree) |> formatC(big.mark = ",")` individual trees found across `r length(unique(tree$PlotID))` plots. @fig-map shows the locations of plots across the pae ʻāina.

```{r}
#| label: fig-map
#| fig-height: 3
#| fig-cap: "Distribution of plot locations across the pae ʻāina Hawaiʻi. Points are semi-transparent to better show plot locations when those locations are very proximate, thus darker colors represent a higher density of plots."

ggplot(hi) +
    geom_sf() +
    geom_sf(data = clim_st, color = hsv(0.65, 0.8, 0.5), alpha = 0.25) +
    theme_void()
```

Rows in the forest plot data represent individual trees. Therefore, some data across the columns will be duplicated, for example trees from the same plot will have the same plot ID. Columns in the forest plot data are described in `data/README_for_OpenNahele_Tree_Data.txt`, reproduced below:

```{r}
#| label: tree-meta

tmeta <- read.table("data/README_for_OpenNahele_Tree_Data.txt", 
                    sep = "\t", 
                    header = TRUE)
names(tmeta) <- gsub("\\.", " ", names(tmeta))

kable(tmeta)

```

### Climate data

Climate data are found in `data/plot_climate.csv`. Rows here are unique plots and no information in any column is duplicated. Columns are described in `data/README_for_plot_climate.txt`, reproduced below:

```{r}
#| label: clim-meta

cmeta <- read.table("data/README_for_plot_climate.txt", 
                    sep = "\t", 
                    header = TRUE)
names(cmeta) <- gsub("\\.", " ", names(cmeta))

kable(cmeta)
```

### Human impact and geophysical data

Human impact data and geophysical data are found in the same file, only because they are from the same source. That file is `data/hii_geo.csv`. Its rows are also unique plots and no information in columns is duplicated across rows. Columns are described in `data/README_for_hii_geo.txt`, reproduced below:

```{r}
gmeta <- read.table("data/README_for_hii_geo.txt", 
                    sep = "\t", 
                    header = TRUE)
names(gmeta) <- gsub("\\.", " ", names(gmeta))

kable(gmeta)

```

### Biocultural dimensions

```{r}
laau <- read.csv("data_raw/ethnobotany_bishop.csv") |> 
    rename(name = Hawaiian.Name, 
           latin_name = Scientific.Name, 
           haole_name = Vernacular.Name) |> 
    mutate(latin_name = gsub(" \\(.*", "", latin_name))

in_laau <- select(tree, Scientific_name) |> 
    mutate(genus = gsub(" .*", "", Scientific_name)) |> 
    group_by(Scientific_name) |> 
    mutate(in_laau = paste(first(Scientific_name), first(genus), sep = "|") |> 
               grepl(laau$latin_name) |> 
               any()) |> 
    ungroup()

prop_stem_in_laau <- mean(in_laau$in_laau)
prop_spp_in_laau <- mean(in_laau$in_laau[!duplicated(in_laau)])
```

[Lāʻau](https://wehe.hilo.hawaii.edu/?q=laau#w2w2-845) are found in the Kumulipo, an ʻŌiwi creation chant, and many significant genealogical [moʻolelo](https://wehe.hilo.hawaii.edu/?q=moolelo#w2w2-14639) such as that of Hāloanakalaukapalili (the first kalo) and Hāloa (his younger brother, the first Hawaiian), bringing the plants we'll work with into the [moʻokūʻauhau](https://wehe.hilo.hawaii.edu/?q=mookuauhau#w2w2-28752) of Kānaka ʻŌiwi. The author (me Andy your instructor) is not trained to understand these relationships but follows the lead of ʻŌiwi scholars and their work such as @hutchins2023 and @goldbergsilva2015. Genealogical relations require deep regard and respect.

Western science might use the term "biocultural" for the connections between human and non-human aspects of the world. In addition to "biocultural"[^1] moʻokūʻauhau, there are at least two other "biocultural" dimensions to consider. 

[^1]: "Biocultural" is in quotes here to indicate the deficiency of the term in this context and especially when grammatically attached to moʻokūʻauhau

- [Lāʻau lapaʻau](https://wehe.hilo.hawaii.edu/?q=laau%20lapaau#w2w2-37956) and ethnobotanical connections to plants
- [Wahi pana](https://wehe.hilo.hawaii.edu/?q=wahi%20pana#w2w2-33435) and different categories of wao (e.g. [wao akua](https://wehe.hilo.hawaii.edu/?q=wao%20akua#w2w2-72767)) along with other ʻōiwi geographies that carry specific cultural protocols and sensitivities


In regard to lāʻau lapaʻau and ethnobotany, the Bernice Pauahi Bishop Museum provides a [Hawaiian ethnobotany database](https://data.bishopmuseum.org/ethnobotanydb/ethnobotany.php?b=list&o=1) built largely on the work of esteemed ʻŌiwi scholar Isabella Aiona Abbott [-@abbott1992]. This database is neither CARE nor FAIR compliant (not even providing full bibliographic information to fully credit @abbott1992) so we will not linger on it except to compare the plant names listed as ethnobotanically significant with plant names in the data we will work with. Doing so reveals that in the data set we will work with, `r round(prop_stem_in_laau * 100)`% of individual trees and `r round(prop_spp_in_laau * 100)`% of unique taxa have ethnobotanical significance.

```{r}
prop_native_plot <- group_by(tree, PlotID) |> 
    summarize(prop_native = mean(Native_Status == "native"))

prcnt_native <- round(mean(prop_native_plot$prop_native >= 0.75) * 100)
```

A similar analysis of the geographic overlap of wahi pana and wao with the coordinates of plot data is not as simply completed. Different wao have specific geographic boundaries and carry different cultural protocols, knowledge held by cultural practitioners and not always interoperated with western cartography [which would allow analysis with plot data; but see @winter2017]. However, there is a working assumption among some cultural and natural resource practitioners in Hawaiʻi that the wao akua corresponds roughly to high elevation, native forest areas. This is not universally agreed upon but in the current context could have utility as a heuristic. What constitutes native forest is open for debate, and any choice comes with inaccuracies. However, we could ask what percent of plots are composed of at least three quarters native trees. The answer is `r prcnt_native`%. So *perhaps* `r prcnt_native`% of plots are located in wao akua across the pae ʻāina. The wao akua is a realm with strict cultural protocols relating to its [kapu](https://wehe.hilo.hawaii.edu/?q=kapu#w2w2-20310) status [@winter2018], protocols that were likely rarely if ever met when collecting these data. 

Many---perhaps most---wahi pana are recorded in written Hawaiian language sources (e.g. nūpepa) that are cataloged in the [Papakilo Database](https://www.papakilodatabase.com/). However, Andy is not proficient enough in the Papakilo Database nor ʻōlelo Hawaiʻi nor place names across the entire pae ʻāina to analyze plot coordinates in relation to wahi pana, at least for now. But we should assume that many plots are located in or near wahi pana.



## Ecological questions we'll consider

The flora of Hawaiʻi has been a source of inspiration, sustenance, medicine, and scientific inquiry for millennia [@abbott1992]. Kānaka ʻŌiwi produced, shared, and continue to produce the ancestral knowledge needed to understand the flora of Hawaiʻi and its connect to humans. ʻŌiwi scientists do so using ʻŌiwi methodologies [for one compilation see @thapar2013] grounded in epistemologies that at times share parallels with western epistemologies [@johnson1981 highlights parallels to Darwin's theory of natural selection as emphasized by @keawe2008] and at other times do not [@handy-pukui]. ʻŌiwi methodologies are self-evidently rigorous and deserve equal respect with western methodologies. Using ancestral knowledge and Indigenous methodologies, Kānaka ʻŌiwi sustained large populations over long periods of time with effectively zero external resource input, in contrast to the current situation in Hawaiʻi of an unsustainable population, not much larger than pre-contact ʻŌiwi populations, dependent on external inputs for 90% of our needs [@gon2018]. ʻŌiwi methodologies should be learned but this class could not possibly do that justice. Therefore this class stays within western epistemologies, not out of a sense of superiority, but of positionality.

From a western perspective, forests in Hawaiʻi are shaped by a multitude of processes [@barton2021]. This is not a forest ecology class, but it is a class about using statistical models to help us answer questions in ecology and evolution. So we will engage with questions and hypotheses in ecology and evolution. We will foreshadow some of our future analyses by highlighting a few interesting drivers of and processes underlying plant biodiversity in Hawaiʻi. 

Two striking features about the flora of Hawaiʻi are 1) it is abundant with diverse and unique forms of plant life; and 2) it is highly modified by human activities, notably pervasive and numerous invasive or non-native species brought by settler colonialism. Many students may not be plant people, [e kala mai](https://wehe.hilo.hawaii.edu/?q=kala#w2w2-11352), so it is worth saying that neither, really, is Andy, and the two striking features about the flora of Hawaiʻi are true for the larger *biota* of Hawaiʻi. While the unique physiological and demographic mechanisms underlying the flora are surely different from the fauna, the courser ecological and evolutionary processes likely share some generalities.

Of those possible generalities, we will look into a couple---with statistical models---throughout this class:

- How have ecology and evolution played out across the multi-million year chronosequence of the pae ʻāina to shape differences in biodiversity across and within islands [investigated, for example, by @gillespie2016; @rominger2016]?
- How have human impacts and climatic variables modulated biodiversity [investigated, for example, by @lim2021; @hutchins2023]
- Can data on human impacts and climate help us understand how invasive/non-native species occupy ecosystems [investigated, for example, by @blackburn2011; @fortini2013]


## Slides {#sec-slides}

```{=html}
<iframe class="slide-deck" src="data-slides.html"></iframe>
```


## References


