---
title: "Null hypothesis testing considerations and connections"
---

```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false

library(knitr)
library(ggplot2)
library(cowplot)
library(dplyr)
library(MASS)

opts_chunk$set(echo = TRUE)

# shared data setup used throughout
clim <- read.csv("data/plot_climate.csv")
geo_hii <- read.csv("data/hii_geo.csv")
tree <- read.csv("data/OpenNahele_Tree_Data.csv")

all_dat <- left_join(tree, clim, by = "PlotID") |>
    left_join(dplyr::select(geo_hii, !(lon:lat)), by = "PlotID")

dat_plt <- group_by(all_dat, PlotID) |>
    summarize(across(c(Island, Plot_Area, lon:elev_m), first),
              nspp = n_distinct(Scientific_name)) |>
    ungroup() |>
    filter(Plot_Area < 10000)
```


## The GLM

We've been working with models like

```{r}
sar_mod <- glm.nb(nspp ~ log(Plot_Area), data = dat_plt)

```

::: {.fragment}

<br>
And testing "significance" with a likelihood ratio test (LRT)

:::


::: {.fragment}

<br>
But there are other approaches, perhaps some that seem more familiar

:::

## The GLM {.scrollable}

The `summary` function gives us a lot of info (scroll down)

```{r}
summary(sar_mod)
```


## The GLM 

Some of this info in fact directly relates to the LRT

```{r}
#| echo: false

s <- capture.output(summary(sar_mod))

s[15:16]
```


::: {.fragment}

(`527` d.f.) - (`526` d.f.) = 1 d.f., the right number if we were comparing models

```{r}
#| eval: false

mod_full <- nspp ~ log(Plot_Area)
mod_reduced <- nspp ~ 1
```

:::


::: {.fragment}

The LRT statistic for those two models is

```{r}
#| echo: false
mod_red <- glm.nb(nspp ~ 1, data = dat_plt)

lam <- 2 * (logLik(sar_mod) - logLik(mod_red))
```

$$
\begin{aligned}
\Lambda &= 2 (\log\mathscr{L}_\text{full} - \log\mathscr{L}_\text{reduced}) \\
&= `r round(as.numeric(lam), 2)`
\end{aligned}
$$

which is (approx.) equal to `622.95 - 515.85`

(approx. because printed summary for `glm.nb` is a little strange)

:::



## The GLM 

What's going on?

::: {.fragment}

The *deviance* is an attempt to generalize the idea of *"sum of squares"*

:::


::: {.fragment}

$$
\begin{aligned}
\text{Null deviance} &= 2 (\log\mathscr{L}_\text{saturated} - \log\mathscr{L}_\text{reduced}) \\
\text{Residual deviance} &= 2 (\log\mathscr{L}_\text{saturated} - \log\mathscr{L}_\text{full})
\end{aligned}
$$
Where $\log\mathscr{L}_\text{saturated}$ is the log likelihood of a model where every data point gets a parameter (i.e. its the highest possible log likelihood)
:::


## The GLM 

What's going on?

::: {.columns}


:::: {.column width=30%}

*Null deviance* 

<br>

*Residual deviance*

::::

:::: {.column width=70%}


corresponds to "total sum of squares" in a least squares world


corresponds to "residual sum of squares"

::::


:::

::: {.fragment}

In least squares world we do an $F$-test (aka ANOVA) which compares residual and total sums of squares

:::

::: {.fragment}

<br>
In likelihood world we do an LRT (perhaps with function `anova`) which compares residual and null deviances

:::


## The GLM versus the LM

The only difference between a GLM and LM is an LM assumes normality


::: {.fragment}

It is a mathematical novelty that under normality, 


::: {.columns}

:::: {.column width=45%}

::: {style="text-align: right;"}

least squares 

minimize sum of squares residual 

$P$-value of SumSq ANOVA

:::

::::

:::: {.column width=10%}

::: {style="text-align: center;"}

$\Longleftrightarrow$

$\Longleftrightarrow$

<br>
$=$

::: 

::::

:::: {.column width=45%}

maximum likelihood

maximize log likelihood function

$P$-value of LRT

::::

:::

:::


## The GLM 

The `summary` function also gave $P$-values

```{r}
#| echo: false

s[6:11]
```

What are those about?

::: {.fragment .incremental}

- Those come from the asymptotic (CLT) behavior of maximum likelihood estimates being normally distributed
- They are generally not as robust as the $t$-test we can do for LM with normality assumption
- The LRT is preferred for significance testing

:::


## The mixed effects model

Recall for a mixed effects model we did not get any $P$-values despite `summary` reporting $t$-values ($t$ as in $T$-test)

```{r}
#| echo: false


dat <- left_join(tree, geo_hii) |> 
    # remove missing DBH values
    filter(!is.na(DBH_cm)) |> 
    # add a column for log-transformed DBH since this will be 
    # the variable we use as a response variable
    mutate(log_dbh_cm = log(DBH_cm)) |> 
    # remove NAs for elevation
    filter(!is.na(hii)) |> 
    # remove "uncertain" for native status
    filter(Native_Status != "uncertain")

```

```{r}
library(lme4)

mixed_mod <- lmer(log_dbh_cm ~ Native_Status * hii + (1 | PlotID), 
                  data = dat, REML = FALSE)

summary(mixed_mod)
```


## The mixed effects model

Here again an LRT is the better option

```{r}
# a model to single out the interaction term
reduced_mod <- lmer(log_dbh_cm ~ Native_Status + hii + (1 | PlotID), 
                  data = dat, REML = FALSE)

# the LRT
anova(mixed_mod, reduced_mod, test = "Chisq")
```
