---
format:
  html:
    code-links:
      - text: View source
        icon: github
        href: https://github.com/uhm-biostats/stat-mod/blob/main/lin-norm.qmd
---

# Simple linear models

This lab will revisit simple linear models from a likelihood framework.

You can download the lab template `.qmd` file for this lab here:

{{< downloadthis labs/lab04-simple-linear.qmd dname=lab04-simple-linear label="Download lab template">}}

The template will prompt you to work through the following sections.

```{r}
#| include: false

library(ggplot2)
library(dplyr)

# the lab qmd file
f <- "labs/lab04-simple-linear.qmd"
l <- readLines(f)

# the yaml
iyaml <- grep("---|```", l) |> 
    (\(x) x[1]:x[4])()

# the comments to delete
idelt <- grep("delete .* once complete", l, 
              ignore.case = TRUE)

# only the body
b <- l[-c(iyaml, idelt)] |> 
    gsub("\\*delete .* once complete", "", x = _, 
         ignore.case = TRUE)

# write only the body for use with `{{< include %s >}}`
# fb <- gsub("\\.qmd", "_bonly.qmd", f)
fb <- tempfile(fileext = ".qmd")

writeLines(b, fb)

m <- knitr::knit_child(fb, options = list(include = TRUE), 
                       quiet = TRUE)
```


```{r}
#| echo: false
#| results: asis

cat(m, sep = "\n")
```

