---
title: "Ideas on how to use large language models"
---

```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false

library(knitr)
opts_chunk$set(echo = FALSE)
```



## Never copy data into an LLM

- Ethical considerations: are the data yours to share?
- Practical considerations: uses up unnecessary LLM memory and context (taking away from a better answer)
- Learning considerations: you'll learn better if you think more critically about what your data are and what questions you really need an LLM's help with

::: {.fragment}
On that note...
:::

## Use hypotheticals

For example, your prompt could be

> I have a \`data.frame\` with columns \`x\` and \`y\`, <br> how do I....

::: {.fragment}
Most LLMs will easily work with that prompt to first generate a toy data object like

```{r}
#| eval: false
#| echo: true

my_data <- data.frame(x = rnorm(10), y = rnorm(10))

```

And then address your question
:::

## Use hypotheticals {auto-animate="true"}

This is useful because

- it makes you think about your data and what you need to figure out
- it keeps the real data private
- having the LLM make up toy data gives you a chance to evaluate the LLM to see if it's really understanding your context and question


## Break your problem into smaller parts

Rather than prompting

> I want to model how \`y\` changes with \`x\` based on group \`g\`, how?

::: {.fragment}
Try breaking it up with your human brain first and figure out what steps you actually want to ask about

1. Prep data for modeling (suppose you already know how)
2. Identify the model formula: `y ~ x * g` (suppose you already know how)
3. Use an R function to fit the model (perhaps this is where you are unsure)
:::

## Break your problem into smaller parts {auto-animate="true"}

Rather than prompting

> I want to model how \`y\` changes with \`x\` based on group \`g\`, how?


Ask specifically about what you don't know!

::: {.fragment}
> I have data with columns \`y\`, \`x\`, and \`g\`. In R I want to model \`y ~ x * g\` but not sure which function to use, how could I figure that out?
:::

## Push back on the LLM

- if the answer is too broad, further refine your prompt to focus the LLM, think back to breaking up the problem into smaller steps and identifying which you need help with

::: {.fragment}
> I have data with columns \`y\`, \`x\`, and \`g\` where \`x\` and \`y\` are numerical and \`g\` represents different groups. In R I want to model \`y ~ x * g\` but not sure which function to use, how could I figure that out?
:::

## Push back on the LLM {auto-animate="true"}


- if the answer seems unnecessarily complex, ask it, "is there a simpler way to do this?"
- LLMs can really over think things using the tidyverse so ask about non-tidyverse solutions
- ask the LLM to explain steps that seem opaque or that you simply don't understand


## Check the LLM's answers

- run the code yourself, if there are errors, tell the LLM (paste the error message in) and make it fix the mistake
- try asking questions about steps you are fairly sure you already know, if the LLM agrees this will help you know it is headed in the right direction



## Never copy paste code from the LLM and call it done

- make sure you understand every line
- re-write the comments in your own words
- you should have used hypotheticals and not the real data, so at a minimum you will need to change variable names

## Ask for explanations

- show the LLM code or ideas you don't understand and ask it to explain
- topics in this course have a lot of material written about them, so LLMs should be well trained
- try questions like these

## Ask for explanations {auto-animate="true"}

> what is this code doing
>
> \`\`\` <br>
> # insert code here <br>
> \`\`\`

## Ask for explanations {auto-animate="true"}

> why would I need a random effect for data like `_blank_` and a model like `_blank_`?

## Ask for explanations {auto-animate="true"}

> what is different about the `lm` function versus the `glm` function?



## The "what if..."

What if you don't follow these guides?

- First off, we're all learning LLMs together, so you might have your own guides that work just fine
- But second, if you don't follow these guides nothing will blow up...not right away anyway
- The concern is that without understanding the code and concepts you could introduce errors into your analyses
- Sometimes such errors could have serious consequences
