---
title: "Likelihood Ratio Tests and Confidence Intervals"
---

```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false

library(knitr)
library(ggplot2)
library(cowplot)
library(dplyr)
library(MASS)

opts_chunk$set(echo = FALSE)

# shared data setup used throughout
clim <- read.csv("data/plot_climate.csv")
geo_hii <- read.csv("data/hii_geo.csv")
tree <- read.csv("data/OpenNahele_Tree_Data.csv")

all_dat <- left_join(tree, clim, by = "PlotID") |>
    left_join(dplyr::select(geo_hii, !(lon:lat)), by = "PlotID")

dat_plt <- group_by(all_dat, PlotID) |>
    summarize(across(c(Island, Plot_Area, lon:elev_m), first),
              nspp = n_distinct(Scientific_name)) |>
    ungroup() |>
    filter(Plot_Area < 10000)
```


## Is overdispersion supported by my data?

We saw that the Poisson GLM assumes variance = mean = $\lambda$

::: {.fragment}

A negative binomial GLM relaxes that: variance $= \lambda + \lambda^2 / k$, where $k$ is an extra dispersion parameter

:::

::: {.fragment}

```{r}
#| echo: true

mod_pois <- glm(nspp ~ log(Plot_Area),
                data = dat_plt, family = poisson)

mod_nb <- glm.nb(nspp ~ log(Plot_Area),
                 data = dat_plt)
```

Both models fit. But is the negative binomial actually *better*?

:::


## Is overdispersion supported by my data?

We can compare log likelihoods

```{r}
#| echo: true

logLik(mod_pois)
logLik(mod_nb)
```

::: {.fragment}

The negative binomial has a higher log likelihood, but is that difference *significant*?

:::


## What to consider for significance

::: {.incremental}

1. **Quantify** the difference between models
2. **Compare** that quantity to what we would expect if there were *no* real difference, i.e. the null

:::




## What is the null?

If the two models were truly equivalent (no overdispersion, $k \rightarrow \infty$), what would the difference in log likelihoods look like?

::: {.fragment}

- Adding a parameter *always* increases log likelihood a little, even when fitting noise
- If the null is true, the gain is just random luck

:::

::: {.fragment}

We need a null distribution for accidental log likelihood gains

:::


## What is the null?

Need to consider: more parameters let the model bend to match idiosyncrasies of the data, even pure noise (the case here)

```{r}
#| fig-width: 10
#| fig-height: 4
#| fig-align: center

set.seed(7)
n_toy <- 12
x_toy <- seq(0, 1, length.out = n_toy)
y_toy <- rnorm(n_toy, mean = 2, sd = 0.6)

fit_line <- lm(y_toy ~ poly(x_toy, 1, raw = TRUE))
fit_quad <- lm(y_toy ~ poly(x_toy, 3, raw = TRUE))
fit_overfit <- lm(y_toy ~ poly(x_toy, 9, raw = TRUE))

x_seq <- seq(0, 1, length.out = 300)
pred_df <- data.frame(
    x    = rep(x_seq, 3),
    y    = c(predict(fit_line,    newdata = data.frame(x_toy = x_seq)),
             predict(fit_quad,    newdata = data.frame(x_toy = x_seq)),
             predict(fit_overfit, newdata = data.frame(x_toy = x_seq))),
    model = rep(c("2 parameters",
                  "4 parameters",
                  "10 parameters"), each = 300)
)
pred_df$model <- factor(pred_df$model,
                        levels = c("2 parameters",
                                   "4 parameters",
                                   "10 parameters"))

dat_toy <- data.frame(x = x_toy, y = y_toy)

ll_labels <- data.frame(
    model = factor(c("2 parameters",
                     "4 parameters",
                     "10 parameters"),
                   levels = c("2 parameters",
                              "4 parameters",
                              "10 parameters")),
    label = sprintf("logLik = %.1f",
                    c(logLik(fit_line),
                      logLik(fit_quad),
                      logLik(fit_overfit)))
)

ggplot(pred_df, aes(x, y)) +
    geom_line(linewidth = 1.5, color = hsv(0.52, 1, 0.7)) +
    geom_point(data = dat_toy, aes(x, y), size = 3, inherit.aes = FALSE) +
    geom_label(data = ll_labels, aes(label = label),
               x = 0.5, y = 0.5,
               size = 4.5, inherit.aes = FALSE) +
    facet_wrap(vars(model)) +
    xlab("x") +
    ylab("y") +
    theme_cowplot() +
    theme(axis.text = element_blank(),
          axis.ticks = element_blank(),
          axis.title = element_text(size = 20),
          strip.text = element_text(size = 16))
```



## What is the null?

How do more parameters increase likelihood? 

Access to higher parameter dimensions.

::: {.columns}

:::: {.column width=60%}

```{r}
#| fig-width: 6
#| fig-height: 6
#| fig-align: center


p0 <- 0.0   # peak location along b0 axis
p1 <- 0.0   # peak location along b1 axis
curv <- 1.0   # equal curvature in both directions -> circular rim
depth <- 4.0   # how far the bowl drops: peak_ll - depth = floor
peak_ll <- 0     # value at the peak

ll_bowl <- function(b0, b1) -(curv * (b0 - p0)^2 + curv * (b1 - p1)^2) + peak_ll

# rim radius: where ll_bowl == peak_ll - depth
# curv * r^2 = depth  =>  r = sqrt(depth / curv)
rim_r <- sqrt(depth / curv)

# grid just wide enough to contain the circle
b0_vec <- seq(p0 - rim_r, p0 + rim_r, length.out = 80)
b1_vec <- seq(p1 - rim_r, p1 + rim_r, length.out = 80)
ll_mat <- outer(b0_vec, b1_vec, ll_bowl)

# clip to NA outside the circle so the rim sits exactly on the floor
dist_mat <- outer(b0_vec - p0, b1_vec - p1, function(d0, d1) sqrt(d0^2 + d1^2))
ll_mat[dist_mat > rim_r] <- NA

# color each facet by height
n_col <- 100
col_ramp <- hcl.colors(n_col, palette = "viridis")
facet_z <- (ll_mat[-1, -1] + ll_mat[-1, -ncol(ll_mat)] +
                ll_mat[-nrow(ll_mat), -1] + 
                ll_mat[-nrow(ll_mat), -ncol(ll_mat)]) / 4
facet_col <- col_ramp[cut(facet_z, breaks = n_col)]


# 1D parabolic slice: fix b0 = 1, vary b1
# clip to NA where the slice falls outside the bowl (dist > rim_r)
b1_curve <- seq(p1 - rim_r, p1 + rim_r, length.out = 300)
b0_curve <- rep(1, 300)
ll_curve <- ll_bowl(b0_curve, b1_curve)

# any point where distance from bowl center > rim_r is outside — set to NA
outside <- sqrt((b0_curve - p0)^2 + (b1_curve - p1)^2) > rim_r
ll_curve[outside] <- NA


# peak-to-peak connector
b0_connect <- seq(1, p0, length.out = 100)
b1_connect <- rep(0, 100)
ll_connect <- ll_bowl(b0_connect, b1_connect)


# plot
par(mar = c(2, 1, 0, 1), cex = 2)
pmat <- persp(
    x = b0_vec, y = b1_vec, z = ll_mat,
    col = facet_col,
    border = "transparent",
    theta = 60,
    phi = 22,
    expand = 0.6,
    xlab = "b0", ylab = "b1", zlab = "log likelihood",
    shade = 0.3
)

# 1D parabolic slice (clipped to bowl interior)
curve_proj <- trans3d(b0_curve, b1_curve, ll_curve, pmat)
lines(curve_proj, col = "gray", lwd = 4)

# peak-to-peak connector along the bowl surface
connect_proj <- trans3d(b0_connect, b1_connect, ll_connect, pmat)
lines(connect_proj, col = "black", lwd = 4)

# mark bowl peak
peak_proj <- trans3d(p0, p1, peak_ll, pmat)
points(peak_proj, col = "black", pch = 16, cex = 2)

# mark parabola peak (b0=1, b1=0)
par_peak_proj <- trans3d(1, 0, ll_bowl(1, 0), pmat)
points(par_peak_proj, col = "gray", pch = 16, cex = 2)

trans3d(cos(seq(-1.5, 1, length.out = 100)) * rim_r, 
       sin(seq(-1.5, 1, length.out = 100)) * rim_r, 
       rep(-3.8, 10), 
       pmat) |> 
    lines(col = "white", lwd = 10)

```

::::

:::: {.column width=40%}

Log likelihoods are approximately **quadratic**

::: {.fragment}

Thus vertical distances are **squared**

:::

::::

:::


## What is the null?

If vertical distances are **squared**

$2(\log\mathscr{L}(\text{data} \mid b_0,~b_1) - \log\mathscr{L}(\text{data} \mid b_0)) \sim \chi^2$

::: {.columns}

:::: {.column width=60%}

```{r}
#| fig-width: 6
#| fig-height: 6
#| fig-align: center


p0 <- 0.0   # peak location along b0 axis
p1 <- 0.0   # peak location along b1 axis
curv <- 1.0   # equal curvature in both directions -> circular rim
depth <- 4.0   # how far the bowl drops: peak_ll - depth = floor
peak_ll <- 0     # value at the peak

ll_bowl <- function(b0, b1) -(curv * (b0 - p0)^2 + curv * (b1 - p1)^2) + peak_ll

# rim radius: where ll_bowl == peak_ll - depth
# curv * r^2 = depth  =>  r = sqrt(depth / curv)
rim_r <- sqrt(depth / curv)

# grid just wide enough to contain the circle
b0_vec <- seq(p0 - rim_r, p0 + rim_r, length.out = 80)
b1_vec <- seq(p1 - rim_r, p1 + rim_r, length.out = 80)
ll_mat <- outer(b0_vec, b1_vec, ll_bowl)

# clip to NA outside the circle so the rim sits exactly on the floor
dist_mat <- outer(b0_vec - p0, b1_vec - p1, function(d0, d1) sqrt(d0^2 + d1^2))
ll_mat[dist_mat > rim_r] <- NA

# color each facet by height
n_col <- 100
col_ramp <- hcl.colors(n_col, palette = "viridis")
facet_z <- (ll_mat[-1, -1] + ll_mat[-1, -ncol(ll_mat)] +
                ll_mat[-nrow(ll_mat), -1] + 
                ll_mat[-nrow(ll_mat), -ncol(ll_mat)]) / 4
facet_col <- col_ramp[cut(facet_z, breaks = n_col)]


# 1D parabolic slice: fix b0 = 1, vary b1
# clip to NA where the slice falls outside the bowl (dist > rim_r)
b1_curve <- seq(p1 - rim_r, p1 + rim_r, length.out = 300)
b0_curve <- rep(1, 300)
ll_curve <- ll_bowl(b0_curve, b1_curve)

# any point where distance from bowl center > rim_r is outside — set to NA
outside <- sqrt((b0_curve - p0)^2 + (b1_curve - p1)^2) > rim_r
ll_curve[outside] <- NA


# peak-to-peak connector
b0_connect <- seq(1, p0, length.out = 100)
b1_connect <- rep(0, 100)
ll_connect <- ll_bowl(b0_connect, b1_connect)


# plot
par(mar = c(2, 1, 0, 1), cex = 2)
pmat <- persp(
    x = b0_vec, y = b1_vec, z = ll_mat,
    col = facet_col,
    border = "transparent",
    theta = 60,
    phi = 22,
    expand = 0.6,
    xlab = "b0", ylab = "b1", zlab = "log likelihood",
    shade = 0.3
)

# 1D parabolic slice (clipped to bowl interior)
curve_proj <- trans3d(b0_curve, b1_curve, ll_curve, pmat)
lines(curve_proj, col = "gray", lwd = 4)

# peak-to-peak connector along the bowl surface
connect_proj <- trans3d(b0_connect, b1_connect, ll_connect, pmat)
lines(connect_proj, col = "black", lwd = 4)

# mark bowl peak
peak_proj <- trans3d(p0, p1, peak_ll, pmat)
points(peak_proj, col = "black", pch = 16, cex = 2)

# mark parabola peak (b0=1, b1=0)
par_peak_proj <- trans3d(1, 0, ll_bowl(1, 0), pmat)
points(par_peak_proj, col = "gray", pch = 16, cex = 2)

trans3d(cos(seq(-1.5, 1, length.out = 100)) * rim_r, 
       sin(seq(-1.5, 1, length.out = 100)) * rim_r, 
       rep(-3.8, 10), 
       pmat) |> 
    lines(col = "white", lwd = 10)

```

::::

:::: {.column width=40%}

differences in log likelihoods between models (times 2) will be $\chi^2$-distributed

::: {.fragment}

the *times 2* is a scaling factor from deep in the math

:::



::::

:::




## The likelihood ratio test

This $\chi^2$ distribution provides our null, and gives us a significance test!

$$
\Lambda = 2\!\left(\log\mathscr{L}_\text{full} - \log\mathscr{L}_\text{null}\right) \;\sim\; \chi^2_{df}
$$

::: {.incremental}

- $df$ = difference in number of parameters between models
- If the added parameters *are* ***truly*** meaningful, $\Lambda$ should be large and we should reject the null

:::

```{r}
#| include: false
# hidden: keep simulated data needed for the LRT CI slide later
set.seed(42)
n <- 80
x <- runif(n)
y <- rpois(n, exp(0.5 + 1.2 * x))
m <- glm(y ~ x, family = poisson)
b1_hat <- coef(m)[2]
b1_grid <- seq(b1_hat - 0.2, b1_hat + 0.2, length.out = 200)
ll_vals <- sapply(b1_grid, function(b) {
    mu <- exp(coef(m)[1] + b * x)
    sum(dpois(y, mu, log = TRUE))
})
ll_df <- data.frame(b1 = b1_grid, ll = ll_vals)
```


## LRT applied to overdispersion

```{r}
#| echo: true

ll_pois <- logLik(mod_pois)
ll_nb <- logLik(mod_nb)

Lambda <- 2 * (ll_nb - ll_pois)
Lambda

# negative binomial has one extra parameter (k)
pchisq(Lambda, df = 1, lower.tail = FALSE)
```




## LRT applied to overdispersion


We can also use the built-in `anova` function:

```{r}
#| echo: true

anova(mod_pois, mod_nb, test = "Chisq")
```



## LRT: key assumption

**Nested models**

The LRT requires the null model to be a **special case** of the full model

::: {.incremental}

- Poisson is nested in negative binomial ($k \to \infty$ recovers Poisson)
- `glm(y ~ x)` is nested in `glm(y ~ x + z)`
- **Not** nested: Poisson vs. normal; two models with non-overlapping predictors

:::



## LRT for a predictor: does human impact matter?

After accounting for area, does human impact (`hii`) improve the fit?

```{r}
#| echo: true

mod_nb_area <- glm.nb(nspp ~ log(Plot_Area), data = dat_plt)
mod_nb_area_hii <- glm.nb(nspp ~ log(Plot_Area) + hii, data = dat_plt)
```

::: {.fragment}

```{r}
#| echo: true

logLik(mod_nb_area)
logLik(mod_nb_area_hii)
```

:::


## LRT for a predictor: does human impact matter?

```{r}
#| echo: true

anova(mod_nb_area, mod_nb_area_hii, test = "Chisq")
```

::: {.fragment}

`anova` with `test = "Chisq"` does the LRT for us when comparing nested `glm` objects

:::

::: {.fragment}

One extra parameter (`hii`), so $df = 1$

:::


## LRT for a predictor: interpreting the result

```{r}
#| echo: true

coefficients(mod_nb_area_hii)
```

::: {.fragment}

Despite the slope for `hii` being small (and positive), $P$-value is less than 0.05

:::




## Confidence intervals as another way to evaluate significance

A coefficient is judged significant if its confidence interval **does not include zero**

::: {.fragment}

Two ways to build CIs from likelihood:

1. **Likelihood ratio interval** — invert the LRT
2. **Wald interval** — use the CLT approximation directly

:::



## LRT confidence interval

Find all values of $\theta$ such that the LRT would *not* reject the null $H_0: \theta = \theta_0$

$$
2\!\left(\log\mathscr{L}(\hat\theta) - \log\mathscr{L}(\theta_0)\right) \leq 3.84
$$

Rearranging:

$$
\log\mathscr{L}(\theta_0) \geq \log\mathscr{L}(\hat\theta) - 1.92
$$


## LRT confidence interval

```{r}
#| fig-width: 7
#| fig-height: 4
#| fig-align: center

ll_max <- max(ll_vals)
cutoff <- ll_max - qchisq(0.95, df = 1) / 2

ll_in  <- ll_df[ll_df$ll >= cutoff, ]

ggplot(ll_df, aes(b1, ll)) +
    geom_ribbon(data = ll_in,
                aes(ymin = cutoff, ymax = ll),
                fill = hsv(0.52, 0.6, 0.85), alpha = 0.6) +
    geom_line(linewidth = 2) +
    geom_hline(yintercept = cutoff, color = "red", linewidth = 1.5,
               linetype = "dashed") +
    geom_vline(xintercept = b1_hat, color = "orange", linewidth = 1.5) +
    annotate("text", x = b1_hat + 0.05, y = ll_max - 0.25,
             label = "peak", hjust = 0, size = 5) +
    annotate("text", x = b1_hat + 0.2, y = cutoff + 0.15,
             label = "peak - 1.92", hjust = 0, size = 5, color = "red") +
    xlab(expression(hat(beta)[1])) +
    ylab("log likelihood") +
    theme_cowplot() +
    theme(axis.title = element_text(size = 20),
          axis.text = element_text(size = 18))
```


## LRT confidence interval

In R, `confint` on a `glm` uses this profile likelihood approach by default:

```{r}
#| echo: true

confint(mod_nb_area_hii)
```




## Wald confidence intervals

Another approach: use the CLT approximation directly

According to CLT $\hat\theta$ is approximately normally distributed:

$$
\hat\theta \pm z_{0.975} \cdot \widehat{\text{SE}}(\hat\theta)
$$

::: {.fragment}

```{r}
#| echo: true

confint.default(mod_nb_area_hii)
```

`confint.default` uses the Wald approach — fast but less accurate

:::


## LRT vs. Wald intervals

::: {.columns}

:::: {.column}

**LRT (profile likelihood)**

- more accurate, especially for small $n$
- CI can be asymmetric — follows the actual shape of the likelihood surface

::::

:::: {.column}

**Wald**

- fast and simple
- symmetric by construction
- less accurate when log likelihood surface is asymmetric

::::

:::


## Summary

::: {.incremental}

- The **LRT** compares nested models: $\Lambda = 2(\log\mathscr{L}_\text{full} - \log\mathscr{L}_\text{null}) \sim \chi^2_{df}$
- The $\chi^2$ null distribution arises from the **CLT** applied to MLE — the log likelihood surface is approximately quadratic
- **Nested models** are required: null must be a special case of full
- `anova(mod1, mod2, test = "Chisq")` does the work in R

:::

## Summary

::: {.incremental}

- **LRT confidence intervals** invert the LRT: all $\theta$ within 1.92 log likelihood units of the peak form the 95% CI
- **Wald intervals** use the normal approximation directly 

:::
