---
format:
  html:
    code-links:
      - text: View source
        icon: github
        href: https://github.com/uhm-biostats/stat-mod/blob/main/norm-lab.qmd
---

# Estimating the parameters of a normal distribution

This lab will introduce you to another log likelihood function, that of a normal distribution. For data we will investigate annual evapotranspiration (ET). ET is the process of water transferring from plants, soil, and surface water into the atmosphere and plays a central role in governing (or at least *predicting*) ecosystem productivity [@idol2007; @brauman2012] and biodiversity [@fisher2011].

You can download the lab template `.qmd` file for this lab here:

{{< downloadthis labs/lab03-normal-likelihood.qmd dname=lab03-norm-likelihood label="Download lab template">}}

The template will prompt you to work through the following sections.

```{r}
#| include: false

library(ggplot2)

source("R/utils.R")
out <- render_lab("labs/lab03-normal-likelihood.qmd")
```

```{r}
#| echo: false
#| results: asis

cat(out$m, sep = "\n")
```

This same approach will work for calculating the real log likelihood surface.

## References
