# Estimating the parameters of a normal distribution

This lab will introduce you to another log likelihood function, that of a normal distribution. For data we will investigate annual evapotranspiration (ET). ET is the process of water transferring from plants, soil, and surface water into the atmosphere and plays a central role in governing (or at least *predicting*) ecosystem productivity [@idol2007; @brauman2012] and biodiversity [@fisher2011].

You can download the lab template `.qmd` file for this lab here:

{{< downloadthis labs/lab03-normal-likelihood.qmd dname=lab03-norm-likelihood label="Download lab template">}}

The template will prompt you to work through the following sections.

```{r}
#| include: false

library(ggplot2)

# the lab qmd file
f <- "labs/lab03-normal-likelihood.qmd"
l <- readLines(f)

# the yaml
iyaml <- grep("---|```", l) |> 
    (\(x) x[1]:x[4])()

# the comments to delete
idelt <- grep("delete this .* once complete", l, 
              ignore.case = TRUE)

# only the body
b <- l[-c(iyaml, idelt)] |> 
    gsub("\\*delete this .* once complete", "", x = _, 
         ignore.case = TRUE)

# write only the body for use with `{{< include %s >}}`
# fb <- gsub("\\.qmd", "_bonly.qmd", f)
fb <- tempfile(fileext = ".qmd")

writeLines(b, fb)

m <- knitr::knit_child(fb, options = list(include = TRUE), 
                       quiet = TRUE)
```


```{r}
#| echo: false
#| results: asis

cat(m, sep = "\n")
```

This same approach will work for calculating the real log likelihood surface.

## References
