---
title: "The Poisson Generalized Linear Model"

bibliography: stat-mod.bib
csl: conservation-biology.csl
---

```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false

library(knitr)
library(dplyr)
library(ggplot2)
library(cowplot)
library(ggpointdensity)

opts_chunk$set(echo = FALSE)
```


## Consider the Poisson distribution... 

$$
y \sim \text{Pois}(\lambda)
$$

```{r}
#| fig-width: 7
#| fig-height: 4
#| fig-align: center

nn <- 0:12
pois_pmf <- ggplot(data.frame(x = nn, y = dpois(nn, 4)), aes(x, y)) +
    geom_bar(stat = "identity") +
    xlab(expression(italic(y))) +
    ylab("Probability") +
    theme_cowplot() +
    theme(axis.title = element_text(size = 30), 
          axis.text = element_text(size = 24)) +
    geom_vline(xintercept = 4, color = "orange", linewidth = 2)

pois_pmf
```


::: {.absolute top=550 left=0}

::: {.r-stack}

:::: {.fragment .fade-in-then-out}

Often used to model ***counts*** of things (number of trees, number of species, etc.)

::::

:::: {.fragment}

$\lambda$ is both the *mean* and the *variance* of the distribution

::::

:::

:::

## Consider the Poisson distribution... 

$\lambda$ is both the *mean* and the *variance* of the distribution. <br> ***What must be true about*** $\mathbf{\lambda}$?

```{r}
#| fig-width: 7
#| fig-height: 4
#| fig-align: center

pois_pmf
```

::: {.absolute top=600 left=0}

::: {.fragment}

**must be true**: $\lambda \geq 0$

:::

:::

## Consider the Poisson GLM... 

If we wanted to use a Poisson distribution in a GLM, we have to make sure $\lambda \geq 0$

::: {.columns}

:::: {.column width=60%}

```{r}
#| fig-width: 5
#| fig-height: 4
#| fig-align: center

ggplot(data.frame(x = 0:4, y = 0), aes(x, y)) +
    geom_function(fun = \(x) -2 + x, linewidth = 2) +
    ylab(expression(italic(lambda))) +
    theme_cowplot() +
    theme(axis.title = element_text(size = 30), 
          axis.text = element_text(size = 24)) 

```

::::

:::: {.column width=40%}

Can't have this

$$
\lambda = b_0 + b_1 x
$$

Need some kind of link function

::::

:::



## Consider the Poisson GLM... 

If we wanted to use a Poisson distribution in a GLM, we have to make sure $\lambda \geq 0$

::: {.columns}

:::: {.column width=60%}

```{r}
#| warning: false
#| message: false
#| fig-width: 5
#| fig-height: 4
#| fig-align: center

ggplot(data.frame(x = 0:4, y = 0), aes(x, y)) +
    geom_function(fun = \(x) -2 + x, linewidth = 2) +
    ylab(expression(log(italic(lambda)))) +
    theme_cowplot() +
    theme(axis.title = element_text(size = 30), 
          axis.text = element_text(size = 24)) 

```

::::

:::: {.column width=40%}

$e^\text{anything} \geq 0$, always

$$
\lambda = e^{b_0 + b_1 x}
$$

And the inverse gives us back the straight line

$$
\log(\lambda) = b_0 + b_1 x
$$

::::

:::

::: {.fragment}

So with Poisson GLM, a log link function is the default

:::


## Poisson GLM with `glm` function in R

Back to species area

$$
\begin{aligned}
S &= cA^z \\
\log(S) &= \log(c) + z\log(A)
\end{aligned}
$$


```{r}
#| label: data-prep

clim <- read.csv("data/plot_climate.csv")
geo_hii <- read.csv("data/hii_geo.csv")
tree <- read.csv("data/OpenNahele_Tree_Data.csv")

all_dat <- left_join(tree, clim, by = "PlotID") |> 
    left_join(dplyr::select(geo_hii, !(lon:lat)), 
              by = "PlotID")

dat_plt <- group_by(all_dat, PlotID) |> 
    summarize(across(c(Island, Plot_Area, lon:elev_m), first), 
              nspp = n_distinct(Scientific_name)) |> 
    ungroup() |> 
    filter(Plot_Area < 10000)
```

::: {.r-stack}

:::: {.fragment .fade-in-then-out}

```{r}
#| label: sar
#| fig-width: 7
#| fig-height: 4
#| fig-align: center

ggplot(dat_plt, aes(Plot_Area, nspp)) +
    geom_pointdensity(size = 3) +
    scale_color_viridis_c() +
    scale_x_log10(guide = guide_axis_logticks(mid = 1.5, long = 2.5)) +
    scale_y_log10(guide = guide_axis_logticks(mid = 1.5, long = 2.5)) +
    xlab("Plot area") +
    ylab("Number of species") +
    theme_cowplot() +
    theme(axis.title = element_text(size = 20), 
          axis.text = element_text(size = 20))
```

::: {.absolute top=550 left=800}
*(note: 2 super large plots excluded)*

:::

::::


:::: {.fragment}

lends itself to a log link function

$$
\begin{aligned}
\bar{S} &= \lambda \\
\log(A) &= x \\
\Rightarrow \log(\bar{S}) &= b_0 + b_1 \log(A)
\end{aligned}
$$

::::

:::

## Poisson GLM with `glm` function in R

::: {.columns}

:::: {.column width=40%}

with data like these

```{r}
select(dat_plt, Plot_Area, nspp) |> 
    head() |> 
    kable()

```

::::

:::: {.column width=60%}

can fit a Poisson model like this

```{r}
#| echo: true

sar <- glm(nspp ~ log(Plot_Area), 
           data = dat_plt, 
           family = poisson)

```

recall: the link function already takes care of logging the mean species richness, no need to log again

::: {.fragment}

```{r}
#| echo: true

coefficients(sar)
```

:::

::::

:::

## What besides area? 

::: {.incremental}

- We are probably interested in more than area
- Area might even be a *"nuisance variable"*
    - one that we have to account for
    - but we're more interested in other variables
- How about `hii` (human impact) and `avg_temp_annual_c` (temperature)?
- Can!

:::


## Adding more explanatory variables

```{r}
#| echo: true
mod <- glm(nspp ~ log(Plot_Area) + hii + avg_temp_annual_c, 
           data = dat_plt, 
           family = poisson)

coefficients(mod)

```

::: {.fragment}
After accounting for area, species richness *increases* (weakly) with both human impact and temperature

:::

::: {.fragment}

Perhaps non-native plants augment species richness [@sax2003]

:::

::: {.fragment}

Or perhaps lowland (i.e. hot) forests were incredibly diverse [@rock1913] and are now lost to heavy human modification

:::

## Adding more explanatory variables

Beyond biological interpretation, adding more variables requires care

```{r}
poi_ll <- function(formula, data) {
    m <- model.frame(formula, data = data)
    X <- model.matrix(m, data = data)
    y <- model.response(m)
    
    ll <- function(pars, X, y) {
        mu <- exp(X %*% pars)
        mu[!is.finite(mu)] <- .Machine$double.xmax
        dpois(y, mu, log = TRUE) |> 
            sum()
    }
    
    mle <- glm(formula, data = data, family = poisson) |> 
        coefficients()
    
    pars_combo <- expand.grid(
        seq(mle[3] - 2 * mle[3], mle[3] + 2 * mle[3], length.out = 80),
        seq(mle[4] - 2 * mle[4], mle[4] + 2 * mle[4], length.out = 80)
    ) |> 
        as.matrix()
    
    all_ll <- sapply(1:nrow(pars_combo), function(i) {
        ll(c(mle[1:2], pars_combo[i, ]), X, y)
    })
    
    d <- data.frame(pars_combo, all_ll)
    names(d) <- c("x", "y", "loglik") #c(colnames(X), "loglik")
    
    ggplot(d, aes(x, y, z = loglik)) +
        geom_tile(mapping = aes(fill = loglik)) +
        scale_fill_viridis_c() +
        geom_contour(color = "white") +
        scale_x_continuous(expand = FALSE) +
        scale_y_continuous(expand = FALSE)
}
```

::: {.fragment}

This is the likelihood surface for our model, visualizing just the coefficients for human impact and temperature

::: {.columns}

:::: {.column}

```{r}
#| fig-width: 5
#| fig-height: 3.5
#| fig-align: center

ll_ridge <- poi_ll(nspp ~ log(Plot_Area) + hii + avg_temp_annual_c, dat_plt) +
    xlab("slope of human impact") +
    ylab("slope of temperature") +
    theme(axis.title = element_text(size = 20), 
          axis.text = element_text(size = 20), 
          legend.title = element_text(size = 20), 
          legend.text = element_text(size = 15))

ll_ridge
```

::::

:::: {.column}

::: {.r-stack}

:::: {.fragment .fade-in-then-out}

There is a long, flat ridge across the likelihood surface

That's a problem because it is difficult and unreliable to figure out the optimal parameter combination

::::

:::: {.fragment}

why?

::::

:::

::::

:::

:::


## Adding more explanatory variables



::: {.columns}

:::: {.column}

***Collinearity*** creates these kinds of ridges

```{r}
#| fig-width: 5
#| fig-height: 3.5
#| fig-align: center


ll_ridge
```

::::

:::: {.column}


::: {.fragment}

Human impact and temperature are ***correlated***

```{r}
#| fig-width: 5
#| fig-height: 3.5
#| fig-align: center

ggplot(dat_plt, aes(hii, avg_temp_annual_c)) +
    geom_point() +
    xlab("human impact") +
    ylab("temperature") +
    theme_cowplot() +
    theme(axis.title = element_text(size = 20), 
          axis.text = element_text(size = 20))

```

:::

::::

:::

::: {.fragment}

Check for collinearity, a common cut off is $-0.6 < r < 0.6$

:::


```{r}
#| eval: false
plot(dat_plt$hii, dat_plt$avg_temp_annual_c)

explc_var <- group_by(all_dat, PlotID) |> 
    summarize(across(evapotrans_annual_mm:elev_m, first)) |> 
    ungroup()

foo <- filter(tree, Native_Status != "uncertain") |> 
    group_by(Scientific_name) |> 
    summarize(tot_abund = n(), 
              n_plot = n_distinct(PlotID), 
              nat_stat = first(Native_Status)) |> 
    ungroup() |> 
    arrange(desc(n_plot))

ggplot(foo, aes(n_plot, tot_abund)) +
    geom_density2d_filled(data = select(foo, !nat_stat), mapping = aes(n_plot, tot_abund),
                          show.legend = FALSE) +
    # stat_density_2d(aes(fill = after_stat(level)), geom = "raster")  
    # scale_fill_gradient(low = "gray20", high = "gray90") +
    scale_fill_discrete(palette = gray.colors) +
    coord_fixed() +
    scale_x_log10() +
    scale_y_log10() +
    geom_pointdensity() +
    scale_color_viridis_c() +
    facet_wrap(vars(nat_stat)) +
    theme_cowplot()


spp_per_plot <- group_by(tree, PlotID) |> 
    summarize(area = first(Plot_Area),
              nspp = n_distinct(Scientific_name)) |> 
    ungroup()

ggplot(spp_per_plot, aes(area, nspp)) +
    geom_pointdensity() +
    scale_color_viridis_c() +
    scale_x_log10(guide = guide_axis_logticks(mid = 0.75, long = 1.75)) +
    scale_y_log10(guide = guide_axis_logticks(mid = 0.75, long = 1.75)) +
    theme_cowplot()




# 1 Metrosideros polymorpha     14058    434 native  
#  2 Cibotium glaucum             8096    178 native  
#  3 Cheirodendron trigynum       3404    151 native  
#  4 Cibotium menziesii           3436    132 native  
#  5 Psidium cattleianum          1969    124 alien   
#  6 Ilex anomala                  504     85 native  
#  7 Acacia koa                    778     65 native  
#  8 Myrsine lessertiana           217     64 native 



wtf <- group_by(tree, PlotID) |> 
    summarize(area = first(Plot_Area), 
              no_ohia = !any(Scientific_name == sp))

unique(wtf$area[wtf$no_ohia])

spp <- foo$Scientific_name[1:8][-5]

spp_by_plot <- mutate(tree, 
                      PlotID = factor(PlotID), 
                      Scientific_name = factor(Scientific_name)) |> 
    group_by(PlotID, Scientific_name, .drop = FALSE) |> 
    summarize(abund = n()) |> 
    ungroup() |> 
    left_join(
        y = mutate(tree, PlotID = factor(PlotID)) |> 
            group_by(PlotID) |> 
            summarize(area = first(Plot_Area))
    )

pval <- rep(NA, length(spp))
gpp <- vector("list", length(spp))
names(pval) <- names(gpp) <- spp






for(sp in spp) {
    d <- filter(spp_by_plot, Scientific_name == sp & area < 10000)
    gp <- ggplot(d, aes(area, abund)) +
        geom_pointdensity() +
        scale_color_viridis_c() +
        scale_x_log10() +
        coord_transform(y = "sqrt") +
        geom_smooth(method = "glm", method.args = list(family = poisson)) +
        ggtitle(sp)
    
    mp <- glm(abund ~ area, data = d, family = poisson)
    mn <- MASS::glm.nb(abund ~ area, data = d)
    mn1 <- glm(abund ~ area, data = d, family = negative.binomial(1000))
    
    pval[sp] <- pchisq(2 * (logLik(mn) - logLik(mp)), df = 1, lower.tail = FALSE)
    gpp[[sp]] <- gp
}

gpp[[7]]

library(MASS)

```



```{r}
#| eval: false
#| echo: false

library(ggcorrplot)
library(GGally)

too_skew <- function(x) {
    avg <- mean(x, na.rm = TRUE)
    med <- median(x, na.rm = TRUE)
    std <- sd(x, na.rm = TRUE)
    
    if(((med - avg) / std) < -0.1) {
        return(TRUE)
    } else {
        return(FALSE)
    }
}

happy_scat <- function(data, mapping, ...) {
    x <- eval_data_col(data, mapping$x)
    y <- eval_data_col(data, mapping$y)
    xskew <- too_skew(x)
    yskew <- too_skew(y)
    
    # browser()
    
    ggplot(data, mapping) +
        geom_pointdensity() +
        scale_x_continuous(transform= ifelse(xskew, "log10", "identity")) +
        scale_y_continuous(transform= ifelse(yskew, "log10", "identity")) +
        scale_color_viridis_c() +
        theme(legend.position = "none")
}

happy_den <- function(data, mapping, ...) {
    x <- eval_data_col(data, mapping$x)
    xskew <- too_skew(x)
    
    ggplot(data, mapping) +
        geom_density() +
        scale_x_continuous(transform= ifelse(xskew, "log10", "identity"))
}

happy_cor <- function(data, mapping, ...) {
    x <- eval_data_col(data, mapping$x)
    y <- eval_data_col(data, mapping$y)
    
    xskew <- too_skew(x)
    yskew <- too_skew(y)
    
    coord_poss <- function(x) {
        x <- x[!is.na(x)]
        
        min(x) + diff(range(x)) / 2
    }
    
    x_poss <- coord_poss(x)
    y_poss <- coord_poss(y)
    
    corr <- cor(x, y, method = "spearman", use = "complete.obs") |> 
                          round(3)
    
    corr_scale <- scales::col_numeric(c("red", "white", "blue"), domain = -1:1)
    
    # browser()
    ggplot(data, mapping) +
        scale_x_continuous(transform= ifelse(xskew, "log10", "identity")) +
        scale_y_continuous(transform= ifelse(yskew, "log10", "identity")) +
        annotate("text", 
                 x = x_poss, 
                 y = y_poss, 
                 label = corr) +
        theme_minimal() +
        theme(axis.text = element_blank(), panel.grid = element_blank(), 
              panel.background = element_rect(fill = corr_scale(corr)))
}




ggpairs(dplyr::select(explc_var, !PlotID), 
        lower = list(continuous = happy_scat),
        diag = list(continuous = happy_den),
        upper = list(continuous = happy_cor))
```

## References
