---
format:
  html:
    code-links:
      - text: View source
        icon: github
        href: https://github.com/uhm-biostats/stat-mod/blob/main/prob-dist.qmd
---

# Probability distributions in R

This lab will introduce you (or provide a refresher) to working with common probability distributions in R. You can download the template `.qmd` file for this lab here:

{{< downloadthis labs/lab02-prob.qmd dname=lab02-prob label="Download lab template">}}

The template will prompt you to work through the following sections.

```{r}
#| include: false

library(ggplot2)

# the lab qmd file
f <- "labs/lab02-prob.qmd"
l <- readLines(f)

# the yaml
iyaml <- grep("---|```", l) |> 
    (\(x) x[1]:x[4])()

# only the body
b <- l[-iyaml] |> 
    gsub("\\*Delete this .* once complete\\*", "", x = _)

# write only the body for use with `{{< include %s >}}`
# fb <- gsub("\\.qmd", "_bonly.qmd", f)
fb <- tempfile(fileext = ".qmd")

writeLines(b, fb)

m <- knitr::knit_child(fb, options = list(include = TRUE), 
                       quiet = TRUE)
```

```{r}
#| echo: false
#| results: asis

cat(m, sep = "\n")
```


