---
title: "Averaging models with AIC"
format:
    pdf:
        include-in-header:
            text: |
                \usepackage{newunicodechar}
                \newunicodechar{ʻ}{`}
author:
    - your name here
---

```{r}
#| label: setup

# load all necessary packages
library(ggplot2)
library(dplyr)

```


## Overall goal

We want to understand *how* to average models with AIC, and then we also want to see for ourselves *why* model averaging is useful for prediction. So we will build a hand full of models, average them, and then compare the predictive abilities of individual models against the predictive ability of the averaged model. For all these goals we will split our real data in half and use one half for training our models, and the other half for testing them.

We will return to the question of what drives the proportion of non-native trees in a plot---something that would be very useful for land managers to be able to predict!  To model proportions we will again be using our old friend the binomial GLM.

## Preparing the data

We need to *join* the tree survey, climate, and geo/human impact `data.frame`s and then use the split-apply-combine workflow to calculate plot-level data.  The plot-level data we will need are:

- all explanatory variables (hint: use `across`)
- number of non-native species
- number of native species

Because we will ultimately be building models with these data and averaging those models, we need to be very conservative in our handling of missing values. If our explanatory variables have differences in terms of which variables are missing, model averaging will fail. So a final step in our data preparation workflow needs to be `filter`ing out any rows that contain `NA` for the explanatory variables we'd like to use. Plan which explanatory variables you'd like to build models with and then use `filter(!is.na(x1) & !is.na(x2) & !is.na(x3) & ... so on)` for all your intendened explanatory variables.

For ease of working together, let's all name our final, combined data object `dat_non_nat`. 

Once those steps are completed, we'll split our data into training and testing sets.  The key is, we want these training and testing sets to be *random* samples from the data. For random samples, we use the function `sample`.  We are sampling *rows*, so this is how it will work


```{r}
#| eval: false

# be sure to switch `eval` to `true` in this chunk

# size of half the data
n_subsamp <- round(nrow(dat_non_nat) / 2)

# random indexes to be used for sampling
rand_idx <- sample(nrow(dat_non_nat), n_subsamp)

# extract the training data
dat_train <- dat_non_nat[rand_idx, ]

# the test data is everything left
dat_test <- dat_non_nat[-rand_idx, ]
```


## Making models and averaging them

Now let's make three different models based on genuine hypotheses about what processes might affect the proportion of non-native species in a plot. Again, for simplicity of working together, let's name those models `mod1`, `mod2`, and `mod3`, even though each of us might have different explanatory variables in the models.

As a refresher, for making a binomial GLM when we know the number of "successes" and "failures", we use this syntax

```{r}
#| eval: false

# `x` is just a place holder for a real explanatory variable
mod1 <- glm(cbind(n_non, n_nat) ~ x, data = dat_train, 
            family = binomial)

```

Notice that we are using `dat_train`! We are only fitting these models on the training data so we can then see how good (or bad) they predict data they have never seen before---`dat_test`.

Once your three models are built, let's average them. We'll use `model.avg` from the *MuMIn* package:

```{r}
#| eval: false

mod_avg <- model.avg(list(mod1, mod2, mod3))
```


For the sake of really understanding how the model averaging works, let's calculate by hand the Akaike weights and compare to what `model.avg` gave us.  The equation for weights is

$$
w_i = \frac{e^{-\Delta\text{AIC}_i / 2}}{\sum_j e^{-\Delta\text{AIC}_j / 2}}
$$

where $\Delta\text{AIC}_i = \text{AIC}(\text{model}_i) - \min(\text{AIC})$. 

When calculating AIC, we might as well use `AICc` from *MuMIn* so our comparison with the output of `model.avg` will be equivalent.

Next let's use the weights to calculate averaged coefficients for the explanatory variables we used in our models. As an example, if explanatory variable `x` showed up in models `mod1` and `mod2`, its averaged coefficient would be


```{r}
#| eval: false

# assume w1 is weight for mod1, and w2 is weight for mod2
coefficients(mod1)["x"] * w1 + coefficients(mod2)["x"] * w2

```

You only need to do this for *one* of your explanatory variables. Compare to the output of `coefficients(mod_avg)`.


## Testing the models

Finally we get to the *why*. Why is it useful to average models?  Because the predictive ability of an averaged model will be greater than the predictive ability of an individual model on its own.  We'll show this to ourselves by comparing the log likelihood values of the test data set under each our three individual models and the averaged model.

We'll need to calculate AIC by hand for this exercise. But how do we calculate the log likelihood for the *test* data when our model was fit to the *training* data? By using `dbinom`.

```{r}
#| eval: false

# first use `predict` to get the predicted binomial p parameter
# for the test data
p_mod1 <- predict(mod1, newdata = dat_test, type = "response")


# then plug the predicted p into dbinom and sum log probs
ll_mod1 <- dbinom(x = dat_test$n_non, 
       size = dat_test$n_non + dat_test$n_nat, 
       prob = p_mod1, 
       log = TRUE) |> 
    sum()
```


Compare the log likelihood values for `mod1`, `mod2`, `mod3`, and `mod_avg` on `dat_test`. Which has the highest log likelihood?  

Bonus question: why are we comparing log likelihoods even though our models have different numbers of parameters that may or may not be nested?

