---
title: "Estimating the parameters of a normal distribution"
format: 
    pdf: 
        include-in-header:
            text: |
                \usepackage{newunicodechar}
                \newunicodechar{ʻ}{`} 
author: 
    - your name here
---

```{r}
#| label: setup

# load all necessary packages
# ... you may need to add to this! can delete this comment once done
library(ggplot2)
library(dplyr)

```



## Central tendency and variability of annual evapotranspiration

What is the mean annual evapotranspiration across the plots in the *OpenNahele* data and just how variable is ET? Let's read in the climate data and find out.

```{r}
#| label: readin-data

clim <- read.csv("https://raw.githubusercontent.com/uhm-biostats/stat-mod/refs/heads/main/data/plot_climate.csv")
```


To find out the average we can simply use the `mean` function:

```{r}
#| label: mean-ET

mean(clim$evapotrans_annual_mm)
```

For variability, standard deviation is a common measure:

```{r}
#| label: sd-ET

sd(clim$evapotrans_annual_mm)
```

But, as we seem to like to do, let's try to take the hard way to getting those answers. Let's use the method of maximum likelihood to find the mean and standard deviation.  First have a look at the shape of the distribution of the data:

```{r}
#| label: fig-ET-hist
#| fig-width: 4
#| fig-height: 3

ggplot(clim, aes(evapotrans_annual_mm)) +
    geom_histogram()
```

Looks a lot like a bell-shaped curve---like a normal distribution.  Thinking about annual ET, this makes sense too: day to day there are a lot of little processes that add up to produce ET, and on top of that *annual* ET literally adds up the ET of each day. So the Central Limit Theorem likely applies, and we shouldn't be too surprised to find that annual ET looks to follow a normal distribution.

## The log likelihood function for a normal distribution

As discussed in the introductory likelihood [lecture notes](likelihood-lec.qmd) the log likelihood function for data that come from a continuous probability distribution uses the same approach of summing log "probabilities" but now we are actually summing log probability ***densities***

$$
\log\mathscr{L}(\text{parameters, data}) = \sum\log PDF(\text{data} \mid \text{parameters})
$$

So our log likelihood function for the annual ET data will involve summing up the output of `dnorm(..., log = TRUE)`. One aspect that will be new and different from the log likelihood function based in the [binomial distribution lab](binom-lab.qmd) is that we need to estimate *two* parameters for the normal distribution: mean $\hat{\mu}$ and standard deviation $\hat{\sigma}$. 

*You can delete this list once complete, or perhaps make some bullets into subheaders, up to you!*

Your tasks are to

- write an R function to calculate the log likelihood function, it should be able to take arguments for the mean and sd (remember, these parameters need to be passed as together in a single vector) and the data
- visualize the log likelihood *surface* (see below, this will require some new skills)
- use `optim` to find the maximum likelihood estimates for $\hat{\mu}$ and $\hat{\sigma}$
- compare the estimates $\hat{\mu}$ and $\hat{\sigma}$ to the values we calculated with `mean` and `sd`


## Visualizing a surface

Part of our task is to visualize a surface. There are multiple ways to do that, but we use `geom_tile`. Check out how that works

```{r}
#| label: fig-geom_tile-demo

# make some simple data
surf <- data.frame(x = c(1, 2, 1, 2), 
                   y = c(1, 1, 2, 2), 
                   z = c(8, 2, 6, 1))

# visualize the `z` axis as a third dimension coded by color
ggplot(surf, aes(x, y, fill = z)) +
    geom_tile() +
    scale_fill_viridis_c()
```


Look at the data underlying that graph

```{r}
surf

```

The `x` and `y` columns lay out a grid while the `z` column provides the "heights" of the surface at each one of those gridpoints. The `geom_tile` geom just fills the space with squares colored according to the values of `z`.

You can use this for your visualization!

But you will not want to make the xy grid by hand. To make a grid over a range of parameter ($\mu$ and $\sigma$ values) use the `expand.grid` function, for example

```{r}
#| label: expand

params <- expand.grid(mu = seq(650, 850, length.out = 20), 
                      sig = seq(100, 200, length.out = 20))

head(params)
```

Now we have a grid of values in the neighborhood of the results we got from `mean(clim$evapotrans_annual_mm)` and `sd(clim$evapotrans_annual_mm)`. In each column of the output of `expand.grid` there will be 20 unique values because in our call to `seq` we specified `length.out = 20`. You could make that larger or smaller---larger if you feel like you don't have enough tiles for a pretty figure, smaller if (hypothetically) its too big of a computational task for your computer. Our task in this lab is nothing to worry about computationally.

The next thing you need to figure out is how to apply your likelihood function to each parameter combination in each of the rows of `params`. Unfortunately you cannot just plug the columns into your log likelihood function. The reason is that you are (hopefully) passing the entire column of ET data into `dnorm` somewhere in your function. If you also pass in column vectors for the `mean` and `sd` arguments of `dnorm` it will get confused about what you're asking for. 

That means we need to plug in the values for `mean` and `sd` one combination at a time. The way to do that based on what we've learned so far is a for loop!  If you are curious about other ways you'll want to dig deep into functions like `lapply` and `vapply` from base R or `rowwise` for a *dplyr*-flavored approach.

To the for loop! Here is how it will work:

```{r}
#| label: loop-demo

# this is a pretend version of a likelihood function
your_lik_fun <- function(pars, data) {
    pars[1] + pars[2] + sum(data)
} 

# pretend data
your_data <- 1:4

# make an empty vector that will hold the log likelihood values
ll_norm <- numeric(nrow(params)) 

for(i in 1:nrow(params)) {
    pars_i <- c(params$mu[i], params$sig[i])
    ll_norm[i] <- your_lik_fun(pars_i, your_data)
}

# combine params and ll_norm for plotting
parms_ll <- data.frame(params, loglik = ll_norm)

head(parms_ll)
```


