---
title: "Working with mixed effects models in R"
format: 
    pdf: 
        include-in-header:
            text: |
                \usepackage{newunicodechar}
                \newunicodechar{ʻ}{`} 
author: 
    - your name here
---

```{r}
#| label: setup

# load all necessary packages
# ... you may need to add to this! can delete this comment once done
library(ggplot2)
library(dplyr)
library(lme4) # you might need to `install.packages` this one

```



## Using simulated data to understand fitting mixed effects models 

Let's start with some simulated data to both make sure we understand how mixed effects model work (and their assumptions) and how the *lme4* package implements fitting mixed effects models.

First we simulate the data. Imagine we have some response variable `y` and some numeric explanatory variable `x` and we have collected data on these variables from 8 sites (think sampling locations across space) with 20 observations coming from each site.

```{r}
#| label: sim-data

# set-up the simulation

# number of sites (will be rand effect)
nsite <- 8

# number of (pseudo) reps per site
nrep <- 20

# sd of the random effect of site
sig_rand <- 1

# sd of the "residuals"
sig_resid <- 0.25

# overall intercept
a0 <- 8

# slope
a1 <- 2

sim_dat <- data.frame(site = rep(paste0("site_", 1:nsite), each = nrep), 
                      # simulate random effect of site, we would never
                      # actually have this in real data
                      site_rand = rep(rnorm(nsite, mean = 0, sd = sig_rand), 
                                      each = nrep),
                      x = runif(nsite * nrep)) |> 
    # add the response variable
    mutate(y = rnorm(nsite * nrep, 
                     mean = a0 + site_rand + a1 * x, 
                     sd = sig_resid))

head(sim_dat)
```


Let's have a look at these data

```{r}
#| label: fig-sim-viz
#| fig-width: 4
#| fig-height: 3

ggplot(sim_dat, aes(x, y, color = site)) +
    geom_point() +
    scale_color_viridis_d()
```

We can see that the random effect of site bumps the observations up or down but does not change the slope of `y` versus `x`. So the rand effect of site only impacts the intercept (more on that shortly).

This is indeed what we should expect because this is how we simulated the data, we said the mean of `y` is 

```
a0 + site_rand + a1 * x
```

The `site_rand` enters this expression as an additional intercept term along with `a0`.


### Introducing *lme4* and the `lmer` function

Now we are going to try to use maximum likelihood to estimate the values of the parameters `a0`, `a1`, `sig_rand`, and `sig_resid` from the simulated data. We know the "true" values of those parameters, so we can evaluate whether the maximum likelihood method implemented in `lmer` worked.

Here is how to use the `lmer` function:

```{r}
#| label: fit-mod

sim_mod <- lmer(y ~ x + (1 | site), data = sim_dat, REML = FALSE)
```

Notice a few things: 

- The fixed effects part of the model is like what we've seen before: `y ~ x`
- The way a random effect is implemented is through the notation `(1 | site)` which means "let there be an additional intercept term conditional on site"
- We also had to specify `REML = FALSE`, this is because the default is `REML = TRUE` and that default actually does not maximize the likelihood function (what we want) but rather maximizes the "relative" or "residual" likelihood (what we don't want)

Now we can look at the results

```{r}

sim_mod

```

The print-out should show us a standard deviation for random effect `site` that is close to `r sig_rand` and a standard deviation for the residuals that is close to `r sig_resid`. If we want to directly access those values we can like this:

```{r}
# extract the fixed effects
fixef(sim_mod)

# extract the standard deviations of random effects
VarCorr(sim_mod)
```


Something interesting about the way mixed effects models are fit is that the algorithm actually infers the optimal values that should be sampled from the random effect to maximize the likelihood. We can see evidence of that by asking for the "coefficients" of the model

```{r}
coef(sim_mod)

```

We see an intercept for each site. The fact that each site gets its own (randomly sampled) intercept is why this kind mixed effects model is call a "random intercepts" model. 

## Fitting a mixed effects model to real data

Now its your turn to take what we learned with simulated data and fit a random intercepts model to the real data. Let's return to the model of trying to predict log DBH based on human impact and invasive status (with an interaction) but now we'll include a random effect of survey plot. The fixed effects part of the model should be the same as before and look like this:

```
log_dbh_cm ~ Native_Status * hii
```


First load and clean the data as we have before:

```{r}
#| label: readin-data

tree <- read.csv("https://raw.githubusercontent.com/uhm-biostats/stat-mod/refs/heads/main/data/OpenNahele_Tree_Data.csv")

hii_geo <- read.csv("https://raw.githubusercontent.com/uhm-biostats/stat-mod/refs/heads/main/data/hii_geo.csv")

# merge the data into one data.frame, clean, add needed columns
dat <- left_join(tree, hii_geo) |> 
    # remove missing DBH values
    filter(!is.na(DBH_cm)) |> 
    # add a column for log-transformed DBH since this will be 
    # the variable we use as a response variable
    mutate(log_dbh_cm = log(DBH_cm)) |> 
    # remove NAs for elevation
    filter(!is.na(hii)) |> 
    # remove "uncertain" for native status
    filter(Native_Status != "uncertain")

```


Now take it from there!

