---
title: "Probability distributions and simulating data in R"
format: 
    pdf: 
        include-in-header:
            text: |
                \usepackage{newunicodechar}
                \newunicodechar{ʻ}{`} 
author: 
    - your name here
    - your name here
    - etc
---

```{r}
#| label: setup

# load all necessary packages
# ... you may need to add to this! can delete this comment once done
library(ggplot2)

```

## Probability distributions in R

There are a multitude of probability distributions available to use through R, not to mention a whole multitude more via contributed packages. R functions related to distributions and random variables come in four varieties:

- `d` functions (e.g. `dnorm` for the normal distribution) provide access to probability mass functions (PMF) for discrete random variables and probability density functions (PDF) for continuous random variables
- `p` functions (e.g. `pnorm`) provide access to probability calculations over ranges of values
- `q` functions (e.g. `qnorm`) allow us to calculate quantiles of probability distributions 
- `r` functions (e.g. `rnorm`) allow us to draw random numbers from probability distributions

R chooses to encode both PMFs and PDFs as `d` functions so it is on us as users to know whether the output of a `d` function is a true probability (for PMFs) or a probability density (for PDFs). 

The `p` functions either sum up probabilities from PMFs or calculate areas under PDF curves, so in any application we always get a ***p***robability from a `p` function. Technically these `p` functions represent cumulative mass (CMF) or cumulative density functions (CDF) because they calculate how much probability has "accumulated" across given values of the random variable associated with the probability distribution.

With that intro out of the way we will work through specific distributions.


## Binomial

Let's return to the example of the number of dispersal events over a fixed number of time steps from the simplified game version of the Unified Neutral Theory of Biodiversity that we [previously discussed](https://uhm-biostats.github.io/stat-mod/prob-intro.html). For simplicity, let's work with multiples of 10 for probabilities and numbers of time steps: $\mathbb{P}(\text{dispersal}) = $ `m = 0.1` and number of time steps = `N = 10`. What is the probability of observing exactly 1 dispersal event given these conditions?

That is a job for the binomial distribution which R refers to as `binom` with all the distributions functions: `dbinom`, `pbinom`, `qbinom`, `rbinom`. We want a probability of one discrete (think PMF) value, so we can use `dbinom`

```{r}
#| label: first-binom-ex

# setup parameters
m <- 0.1
N <- 10

dbinom(1, size = N, prob = m)
```

Notice how we use `dbinom` (and indeed how we will use any `d` function):

- the first argument is the value of the random variable for which we want to calculate a probability (or density)
- after that come a sequence of arguments whose names correspond to parameters of the probability distribution


With a binomial distribution we often refer to the random variable as the "number of successes," in this case then number of dispersal events is number of successes. The number of observations is often called "number of trials," in this case then number of trials is number of game iterations. 

You'll notice the probability of 1 success in  `r N` trials is `r dbinom(1, N, m)` but we set the probability $\mathbb{P}(\text{dispersal}) = m = `r m`$. Why are they not the same? Or at least not an obvious multiplication or addition of one another? The reason is counting. The probability of exactly 1 success in `r N` trials consists of many ways to arrive at that outcome. If `0` is no success and `1` is success, here are just a few of the ways we can get exactly 1 success:

$$
\begin{align}
&\text{1, `r rep(0, N - 1)`} \\
&\text{0, 1, `r rep(0, N - 2)`} \\
&\text{0, 0, 1, `r rep(0, N - 3)`} \\
&\cdots
\end{align}
$$

The counting gets increasingly complex with higher numbers of successes and trials. Luckily, the `*binom` functions take care of that for us. Let's go over the family.

### `dbinom`

Calculate the probability of 0, 1, 2, and 3 successes out of 10 trials with a probability of success equal to 0.2.


### `pbinom`

Using the same parameters calculate the probability of observing 3 or less successes, i.e., $\mathbb{P}(\text{successes} \leq 3)$. 

First use your calculations in the previous step to arrive at the answer. Next figure out how to use the `pbinom` function to arrive at the same answer.

Finally use the `pbinom` function to calculate the probability of observing *greater than* 3 successes using the same parameters as before. 


### `qbinom`

The `q` function turns this calculation around. Rather than calculate the probability for a number of successes $\leq n$, `qbinom` figures out what value, call it $q$, the random variable needs to take on such that we get back a specific probability $p$: so all `q` functions solve this equation for $q$: $\mathbb{P}(n \leq q) = p$.

Use `qbinom` to figure out what value cuts the binomial distribution, as previously parameterized, in half (i.e. 0.5 probability). 

### `rbinom`

The `r` functions let us draw random numbers from probability distributions. Give it a try with the same binomial distribution we've been working with: draw 20 numbers.

Now let's try to visualize the shape of the binomial PMF but with a histogram of random numbers. We need many more draws than 20 to make the histogram look close to the actual PMF, let's go for 10,000. Keep in mind that to use `ggplot` we need to store the random numbers in a `data.frame` or `tibble`.

## Poisson

The Poisson distribution is very useful in ecology---it is often used to model the probability distribution of counts, for example numbers of individuals or numbers of species. It has only one parameter, the mean value, often called $\lambda$. Let's use `dpois` to get a sense of the shape of the distribution by visualizing the PMF. Because we use `dpois` to calculate the actual probabilities we won't use a histogram but rather a bar plot.

```{r}
#| label: fig-pois-pmf
#| fig-cap: "Example PMF of the Poisson distribution"

# the range of random variable values we'll
# calculate probabilities for
nn <- 0:30

# the Poisson parameter lambda for the mean value
la <- 8

# store range of values and probabilities in data.frame
pois_pmf <- data.frame(n = nn, d = dpois(nn, la))



ggplot(pois_pmf, aes(n, d)) +
    geom_bar(stat = "identity") +
    xlab("Count") +
    ylab("Probability")
```


Let's walk through all the `pois` functions like we did with `binom`.


### `dpois`

Use `dpois` with the same parameters we've been using to calculate the probability of the random variable being equal to each of 7, 8, 9, and 10.


### `ppois`

Use `ppois` to calculate the probability $\mathbb{P}(7 \leq n \leq 10)$ from the Poisson distribution we've been working with. Note: you can confirm you have the right answer using the previous calculation with `dpois`, but make sure you also figure out how to use `ppois`.

### `qpois`

Use `qpois` to find the 25%, 50%, and 75% quantiles of the Poisson distribution with $\lambda = 8$ that we've been working with.

### `rpois`

Make a histogram using `rpois` that should look similar to the PMF in @fig-pois-pmf



## Negative Binomial

The one downside of the Poisson distribution is that with only one parameter $\lambda$, it can have a heard time mimicking real world data. In ecology in particular counts are often "zero-inflated," meaning that compared to the Poisson distribution there are too many 0 counts. One solution is the negative binomial distribution which, in addition to a parameter for the mean, has a parameter $k$ (called `size` in R) that can shift the weight of probabilities more toward 0 and small values. When $k$ is infinite (practically speaking just a *large* number like 10,000 will do) the negative binomial is mathematically equivalent to the Poisson distribution. But any smaller value of $k > 0$ will add extra probability for small values of the random variable.

The mean parameter for the negative binomial is often called $\mu$ and, again, the parameter that shifts the balance toward small values is $k$, though inside R's `nbinom` functions they call it `size`. Let's again walk through all the `nbinom` functions.

### `dnbinom`

To prove the point about zero-inflation, let's compare the negative binomial with mean 8 to the Poisson with mean 8. Specifically let's look at the probability $\mathbb{P}(n = 0)$ for both. We will need to select values for the $k$ = `size` parameter as well, let's try values of `0.1`, `1`, and `2`. Here is how `dnbinom` works for $k = 0.1$:

```{r}
# remember size = k
dnbinom(0, mu = 8, size = 0.1)
```

Now you take it from here!

### `pnbinom`

Use `pnbinom` to calculate the probability $\mathbb{P}(n \leq 4)$ with the three different values of $k$ we've been working with.

### `qnbibom`

Use `qnbibom` to find what value cuts the probability distribution in half for each of the $k$ values we've been working with.

### `rnbinom`

Add to the histogram you made using `rpois` another histogram derived from `rnbinom` with $k$ = `size = 1`. You can use different fills with transparent colors or faceting, depending on which you prefer.

## Normal

The normal distribution is a workhorse of statistics, as we will see. It is often the presumed distribution for any *continuous* random variable. That assumption is not without precedent (as we will see in the next lab). For now we will just get antiquated with the distribution.

A normal distribution has two parameters: again it has a parameter for the mean, often referred to as $\mu$, but simply called `mean` in R's `norm` functions. It also has a parameter for the variance of the distribution $\sigma^2$. We often think of the standard deviation $\sigma$ instead of the variance because it is either to think in terms of linear numbers rather than squared numbers. R's `norm` functions use the standard deviation and call it `sd`.


### `dnorm`

The probability density function (PDF) of a normal random variable is *not* strictly speaking probability. But the shape of the PDF is still helpful for thinking about which values of the random variable are more probable than others. 

Use `dnorm` with `mean = 2` and `sd = 1` to make a visualization of the PDF. Note: we are not working with a PMF, so `geom_bar` is not the correct choice here; instead, try `geom_line`.


### `pnorm`

Find the value that cuts a normal distribution in half when `mean = 2`, `sd = 1` compared to `mean = 2`, `sd = 3`. Do you expect those 50% quantile values to be the same or different?  Now compare the 50% quantile values for normal distributions with `mean = 2`, `sd = 1` versus `mean = 4`, `sd = 1`.

### `rnorm`

Make a histogram using `rnorm` with a `mean` and `sd` of your choosing.


<!-- ## Simulating Binomial data: presence and absence across an environmental gradient -->

<!-- ## Simulating Normal data -->

<!-- ### The Central Limit Theorem -->

<!-- ### Simulating diameter at breast height across different environments and life histories -->

