---
title: "Estimating the parameters of a simple linear model"
format: 
    pdf: 
        include-in-header:
            text: |
                \usepackage{newunicodechar}
                \newunicodechar{ʻ}{`} 
author: 
    - your name here
---

```{r}
#| label: setup

# load all necessary packages
# ... you may need to add to this! can delete this comment once done
library(ggplot2)
library(dplyr)

```

This lab picks up where we left off with the lab in [`lab04-simple-linear.qmd`](lab04-simple-linear.qmd). Again, let's start by reading in the data and wrangling it into a workable format

```{r}
#| label: readin-data

# read-in from github
tree <- read.csv("https://raw.githubusercontent.com/uhm-biostats/stat-mod/refs/heads/main/data/OpenNahele_Tree_Data.csv")

hii_geo <- read.csv("https://raw.githubusercontent.com/uhm-biostats/stat-mod/refs/heads/main/data/hii_geo.csv")

# merge the data into one data.frame and clean
dat <- left_join(tree, hii_geo) |> 
    # remove missing DBH values
    filter(!is.na(DBH_cm)) |> 
    # add a column for log-transformed DBH since this will be 
    # the variable we use as a response variable
    mutate(log_dbh_cm = log(DBH_cm)) |> 
    # remove NAs for elevation
    filter(!is.na(hii)) |> 
    # remove "uncertain" for native status
    filter(Native_Status != "uncertain")

```

## Turns out there's an easier way

Previously we found the parameters of our models by creating a likelihood functions "by hand" and maximizing those functions "by hand" using `optim`. It's probably not surprising that there is an easier way.

Because all our models have been *linear* (meaning the variables are linear combinations of each other) we can find the parameters of our model with the built-in R function `lm`, which stands for "linear model."

### Predicting log DBH from native status

Let's use the model where we predict `log_dbh_cm` based on `Native_Status` to demonstrate `lm`

```{r}
nat_stat_mod <- lm(log_dbh_cm ~ Native_Status, data = dat)
```

Before even looking at the results, let's look at the syntax. We are telling `lm` that we want to make a model where `log_dbh_cm` can change based on the groups in `Native_Status`. The tilde symbol `~` communicates to R that `log_dbh_cm` "is a function of" `Native_Status`, or "is predicted by" `Native_Status`. Note also that we tell `lm` where to find those columns by specifying `data = dat`.

Now have a look at the results

```{r}
nat_stat_mod
```

The "coefficients" are the parameters of our model. But they don't look like the parameters we used when maximizing our boutique likelihood function "by hand." What's going on? The reason is the way we specified our model and the way `lm` specifies the model are different. We created a parameter for the mean of native trees `mu_nat` and the mean of non-native trees `mu_non`. 

R does it different. The `(Intercept)` refers to the mean of `"alien"` trees (recall `Native_Status` contains values of `"alien"` and `"native"`). The coefficient `Native_Statusnative` has information about the mean of native trees, but in a kind of weird way: the value `r nat_stat_mod$coefficients["Native_Statusnative"]` is the *difference* between the mean for `"alien"` and the mean for `"native"`. 

So if you compare these results to what you found when maximizing your boutique likelihood by hand you should have roughly the same number for `mu_non` as `(Intercept)`. And `mu_nat - mu_non` should be roughly the same as `Native_Statusnative` from `lm`. The numbers will only be roughly the same because there is always a small amount of error with `optim` whereas `lm` uses exact equations to find its answers.

Those exact equations are based on minimizing the sum of square error, which is how regression was done for decades because the math (for math nerds) is "simple." The surprising result is that minimizing sum of square error is mathematically the same as maximizing the likelihood function. That's a point worth remembering when we move onto other probability distributions besides the normal.

The `lm` function is a huge help, but we still need to know about the likelihood. Luckily there are built-in functions to help. `logLik` tells us the log likelihood value at its maximum

```{r}
logLik(nat_stat_mod)
```

Compare that result to your results from `optim`, specifically the `value` element of the output from `optim`. The numbers should be *roughly* the same.



### Predicting log DBH from human impact

Now work through those same steps but for the model predicting `log_dbh_cm` from `hii`. The steps are:

1. use `lm` to make a model object
2. compare the coefficients with your results from maximizing your boutique likelihood function; explain for yourself why and how the results are similar or different
3. compute the maximum log likelihood with `logLik`

## More complex models thanks to `lm`

Now because we don't have to do everything by hand we can build more complex models. In the example we've been working with there is an interesting story having to do with how log DBH changes based on native status and human impact acting together. To see that we need to introduce an *interaction* effect in our model. This is how:

```{r}
#| label: nat-hii-mod

# an example of a model with an interaction effect
nat_hii_mod <- lm(log_dbh_cm ~ Native_Status * hii, data = dat)

# have a look
nat_hii_mod
```

There are now four coefficients: `(Intercept)`, `Native_Statusnative`, `hii`, and `Native_Statusnative:hii`. That last coefficient `Native_Statusnative:hii` represents the interaction and here's what it means:

First we need to be clear on the interpretation of `hii`---this is the slope of `log_dbh_cm` changing across `hii`, but ***only*** for the group of trees falling into the `"alien"` category. That means the slope of `log_dbh_cm` changing across `hii` is given by adding the `hii` slope to the coefficient `Native_Statusnative:hii`. So the slope for native trees is

```{r}
nat_hii_mod$coefficients["hii"] + nat_hii_mod$coefficients["Native_Statusnative:hii"]
```

The slope for natives is more steep, meaning the dbh of native trees is more severely decreased by human impact compared to non-native trees. Non-native trees are more resilient to human impact.

An interesting finding indeed!

### Visualizing interactions

To understand what an interaction is, let's visualize how the *mean* or `log_dbh_cm` changes across native statuses and human impact. Here's how.

Use `ggplot` to visualize a function

```{r}
#| label: fig-geom_function

# a function representing an equation
my_fun <- function(x) {
    y <- 30 + 0.5 * x
    
    return(y)
}

# set up a ggplot with our data...
ggplot(dat, aes(hii, log_dbh_cm)) +
    # ... but then plot the function, not the data itself
    geom_function(fun = my_fun)
```

You are not limited to plotting just one function:

```{r}
#| label: fig-many-fun

# a function representing an equation
my_next_fun <- function(x) {
    y <- 1 + 2.5 * x
    
    return(y)
}

# plot multiple functions
ggplot(dat, aes(hii, log_dbh_cm)) +
    geom_function(fun = my_fun, color = "black") +
    geom_function(fun = my_next_fun, color = "red")
```


Now your turn: use the framework to plot two lines: one for native trees, one for non-native trees.  All the coefficients you need can be found in 

```{r}
nat_hii_mod$coefficients
```

Lastly, interpret your results, what do they mean to you?
